### 2016-05-23 Paper/Paper.tex
Updated code, paper files added

% do not change this formatting 
\documentclass[10pt]{article}
\parindent=0pt \parskip=8pt
\textwidth=3in
\hoffset=.3in 
\usepackage{amssymb,amsmath,amsthm,mathrsfs} 
\pagestyle{headings}
\usepackage{amsmath}
\newcommand*{\QEDA}{\hfill\ensuremath{\blacksquare}}%
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{url}
\usepackage{shadethm}
\usepackage{mathtools}
\usepackage{systeme}
\usepackage{float}
\usepackage{mathtools}
\setlength\parindent{24pt}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
\usepackage{caption}
\usepackage{subcaption}
\newtheorem{theorem}{Theorem}
\usepackage{dcolumn}
\newcolumntype{2}{D{.}{}{2.0}}
\usepackage{subfig}

\usepackage[utf8]{inputenc}

\newshadetheorem{thm}{Theorem}
\definecolor{shadethmcolor}{HTML}{EDF8FF}
\definecolor{shaderulecolor}{HTML}{45CFFF}
\setlength{\shadeboxrule}{.4pt}

% replace 'n' with the number of the homework assignment
\title{Analyzing Biological Models with the Wasserstein Distance}

% replace 'your name here' with your name
\author{
  Shane Lubold\\
  \texttt{shane.lubold@asu.edu}\\
  Arizona State University
  \and
  Sebastien Motsch \\
  \texttt{smotsch@asu.edu}\\
  Arizona State University
}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
In recent years a number of mathematical models have been proposed that describe biological systems. One needs a way to compare the accuracy of these methods, in relation to each other and in relation to empirical data. The Wasserstein distance is a new metric that can be used for such a purpose. In one dimension, there exists a closed form expression for the Wasserstein distance, but in two or more dimensions, no such expression exists for most functions. In this paper we present a method for approximating the Wasserstein distance in two or more dimensions. We also consider several biological models and demonstrate how the Wasserstein distance can be used to analyze the effictiveness of such models.
\end{abstract}

\section{Introduction}

  Modeling biological systems has increased in popularity in recent years. From bacteria to swarm bees, a number of models have been proposed to describe each system. But the increase in number of models does not necessary improve the comprehension of the underlying population. One needs to be able to compare there predictions and more important to measure their agreement with experimental data. One way to do this is to use the Wassertstein distance, which allows us to compare discrete and continuous dynamics (i.e.e, systems of ODEs and PDEs). But so far Wasserstein distance has been used mainly as a powerful tool to prove analytic results (i.e. existence/uniqueness of solutions). In this paper we use the Wasserstein distance to compare the accuracy of biological models in relation to empirical data.

The challenge is that the Wasserstein distance, defined as a minimization problem,
can only be computed explicity in one dimension. In two or higher dimensions, we must rely on an approximation. In this paper, we develop a numerical method that approximates the Wasserstein distance through the Simplex method for continuous distributions, and the Hungarian algorithm for discrete distributions (sets of points). Once the numerical methods have been implemented and tested, we will use the Wasserstein distance to perform data-model comparison for several dynamics (e.g. pedestrian dynamics, swarming model). In many studies, data-model comparisons are limited to only visual or qualitative agreement. This framework aims at providing quantitative measurement for the agreement (or discrepancy) between a model and experimental data.

The rest of the paper is organized as follows. In Section \ref{sec: Preliminaries}, we present the general problem, outline the theory and notation, and provide some theoretical results regarding the Wasserstein Distance. In Section \ref{sec: OneD}, we provide the theory and computations of the Wasserstein distance in one dimension. We also consider the case of noisy data and show how this affects the computation of the Wasserstein distance. Section \ref{sec: TwoD} outlines the problem in two dimensions, and shows how the simplex method may be used to compute the Wasserstein distance between continuous and discrete distributions. In Section \ref{sec: BioSystems} we examine how models used to study biological systems compare to empirical data through the use of the Wasserstein distance. Section \ref{sec: Conclusion} provides concluding remarks and points to future work.

\section{Literature Review} 
The optimal transport problem, the problem of moving a mass from one location to another in the most efficient way possible, has been studied for several centuries. One way to phrase the problem is this. Consider a number $n$ of bakeries, each of which makes a certain amount of bread, and we know that there are $m$ restaurants that will consume a fixed amount of bread. Since the amount of bread produced and consumed is fixed for each bakery and restaurant, we can model these quantities as probability measures and create a ``density of production" and a ``density of consumption" on some space, which we can consider to be a metric space with the distance between a bakery and a restaurant as the length of the straight line between them. The optimal transport problem then seeks to move the bread from the bakeries to the restaurants by minimizing the total cost of transportation. For a more complete description of this problem, see Section 3 of \cite{CVillani}.

\noindent Recent work has shown a number of theoretical properties of the Wasserstein distance, including the existence of solutions to the optimal transport problem under certain assumptions, \cite{CVillani}, the convexity of the solution space, \cite{AN}, and the weak convergence of measures, \cite{AN}.

\noindent In recent years there has been a proliferation of biological models concerning a vast array of biological systems.

\section{Preliminaries}
\label{sec: Preliminaries}
Given a Polish (complete, separable, metric) space $(M,d)$, define $\mathcal{P}(M)$ as the set of probability measures on $(M, \mathcal{B}(M))$. If $X, Y$ are two Polish spaces, $T$ is a Borel map, and $\phi \in \mathcal{P}(X)$ is a measure, we can define $T_{\#}\phi \in \mathcal{P}(Y)$, which we call the \textit{pushforward of $\phi$ through T}, as 
\begin{equation}
T_{\#}\phi(A) = \phi(T^{-1}(A)), \ \ \ \ \ \forall A \subseteq Y, \  A \  \text{is Borel}.
\label{eq: TransportDefn}
\end{equation}
The pushforward of $\phi$ through $T$ has the property that
\begin{equation}
\int_Y h \ dT_{\#}u = \int_X h \circ T d\mu,
\label{eq: TransportProperty}
\end{equation}
for every Borel function $h: Y \rightarrow \mathbb{R} \cup \{\pm \infty\}$. Define a Borel cost function $c: X \times Y \rightarrow \mathbb{R} \cup \{\infty\}$. Following the notation of \cite{AN}, we call the following problem formulation the \textit{Monge optimal transport problem}.

\noindent \textbf{Monge Optimal Transport Problem:} Let $v \in \mathcal{P}(X)$, $\mu \in \mathcal{P}(Y).$ We define the $j$th Wasserstein distance between $v$ and $\mu$ as
\begin{equation}
W_j(\mu,\nu) = \inf_{T} \sqrt[\leftroot{-2}\uproot{2}j]{\int_{M}c(x, T(x))^{j} \ d\mu(x)},
\label{eq: CostWassDist}
\end{equation}
where the $\inf$ is taken over all $T$ satisfying (\ref{eq: TransportProperty}). As noted in \cite{AN}, Monge's definition can be ill-posed if no such admissible T exists (if $\mu$ is a Dirac delta and $\nu$ is not) or if the constraint that $T_{\#}\mu = \nu$ is not weakly sequentially closed, w.r.t any reasonable weak topology. To resolve these issues, we also present the formulation attributed to Kantorovich, \cite{AN}.

\noindent \textbf{Kantorovich Optimal Transport Problem:}
\noindent For each $\mu \in \mathcal{P}(X)$ and $\nu \in \mathcal{P}(Y)$, let $\Pi(f,g)$ denote the collection of couplings between $\mu$ and $\nu$. In other words, 
\begin{equation}
\Pi(f, g) = \Big\{\pi: \pi(A \times M) = \mu(A), \ \text{and} \ \pi(A \times M) = \nu(A), \ \text{for each $A \in \mathcal{B}(M)$}  \Big\}.
\label{eq: Couplings}
\end{equation}
We then define the $j$-th Wasserstein distance between $\mu$ and $\nu$ as 
\begin{equation}
W_j(f, g) = \inf_{\pi \in \Pi(f, g)}\sqrt[\leftroot{-2}\uproot{2}j]{\int_{M \times M} d(x,y)^{j} d\pi(x, y)},
\label{eq: CouplingsWassDist}
\end{equation}
where the $\inf$ is taken over all $\pi \in \Pi(f, g)$ that satisfy (\ref{eq: Couplings}). In some cases, $W_p(\mu, \nu) = \infty$, so we will restrict ourselves $W_j(f, g)$ to $\mathcal{P}_j(M) \subset \mathcal{P}(M)$, where $\mathcal{P}_j(M)$ denotes the set of probability measures on M with finite $j$-th moment. In other words, 
\begin{equation}
\mathcal{P}_j(M) = \Big\{f \in \mathcal{P}(M): \int_M d(x, y)^j f(dy) < \infty, \ \text{for some $x \in M$} \Big\}.
\label{eq: jPM}
\end{equation}
As shown in \cite{AN}, $\Pi(f, g)$ is a convex set that contains $f \times g$.  If $c$ is lower semicontinuous and bounded below, then there exists a $\pi \in \Pi(f, g)$ that satisfies (\ref{eq: CouplingsWassDist}). \cite{AN} also proves that if $c$ is continuous and $\mu$ is non atomic, then the values of (\ref{eq: CostWassDist}) and (\ref{eq: CouplingsWassDist}) are equal.
It is also possible to show that $W_{j}$ defines a distance function on $\mathcal{P}_j(M)$. Clearly $W_{j}(\mu, \nu) = W_{j}(\nu, \mu)$ and $W_{j}(\nu, \mu) \geq 0$, with equality only holding in the case that $\mu = \nu$. To prove that $W_{j}(\mu, \nu) \leq W_{j}(\nu, \psi) + W_{j}(\psi, \nu)$ for some measure $\psi$, it's convenient to use the so-called gluing lemma, \cite{CVillani}.

\noindent \textbf{Gluing Lemma:} Let $(X_i, \mu_i), i = 1, 2, 3,$ be Polish probability spaces. If $(X_1, X_2)$ is a coupling of $(\mu_1, \mu_2)$ and $(Y_2, Y_3) $ is a coupling of $(\mu_2, \mu_3)$, then it is possible to construct a triple of random variables $(Z_1, Z_2, Z_3)$ such that $(Z_1, Z_2)$ has the same law as $(X_1, X_2)$ and $(Z_2, Z_3)$ has the same law as $(Y_2, Y_3)$.

\noindent For the remainder of the paper, we restict ourselves to $(M,d) = (\mathbb{R}^n, d)$, with $n \geq 1$ and $d$ being the standard metric on $\mathbb{R}^n$.

\section{One Dimensional Case} \label{sec: OneD}
In one dimension, there exists a closed form solution to computing the Wasserstein distance between two probability density functions. Let $f, g: \mathbb{R} \rightarrow \mathbb{R}$ such that
\begin{equation}
\int_{\mathbb{R}} f \ dx = \int_{\mathbb{R}} g \ dx.
\end{equation}
If we define
\begin{equation}
F(x) = \int_{- \infty}^x f(s) \ ds \ \ \ \ \text{and} \ \ \ \ G(x) = \int_{- \infty}^x g(s) \ ds,
\end{equation}
then the $2$nd Wasserstein distance between $f$ and $g$ is
\begin{equation}
W_2^2(f, g) = {\int_{0}^1 |F^{-1}(y) - G^{-1}(y)|^2 \ dy}.
\label{eq: 1DWD}
\end{equation}
From now on, we will refer to the value of ($\ref{eq: 1DWD}$) as the Wasserstein distance between $f$ and $g$, or simply the WD between $f$ and $g$.
One can think of (\ref{eq: 1DWD}) as moving the $j$-th percentile of $f$ to the $j$-th percentile of $g$ (Figure \ref{fig: 1DWD}).

\begin{figure}[H]
\centering
\includegraphics[scale = .4]{2ndPPercentile.pdf}
\caption{Moving $j$-th percentile (shaded) of $f$ to $j$-th percentile of $g$.}
\label{fig: 1DWD}
\end{figure}
%In some cases, it is possible the data one collects are corrupted by noise, and a relavent question is what are the mean and variance of the resulting approximation of the Wasserstein distance. Assume we observe $\textbf{f}$, $\textbf{g}$\footnote{From here on, boldface variables denote random variables.}, and we want to know the expected value and variance of our estimates of $W_1(f,g)$.
%\begin{equation}
%\textbf{f}(x) = f(x) + v, \ \ \ \textbf{g}(x) = g(x) + v, \ \ \ \ v \sim \mathcal{N}(0, \rho^2).
%\end{equation}
%When $j = 1$, (\ref{eq: Alt1DWD}) becomes
%\begin{equation}
%W_1(f, g) = \int_{-\infty}^\infty |f(x) - g(x)| \ dx.
%\end{equation}
%By linearity, we have that
%\begin{align*}
%\mathbb{E}(W_1(\tilde{f}, \tilde{g})) &= \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f}(x) - \textbf{g}(x)| \ dx \Big) \\  & = 
%\int_{-\infty}^\infty |\mathbb{E}(\textbf{f}(x)) - \mathbb{E}(\textbf{g}(x))| \ dx \\
%&= \int_{-\infty}^\infty |f(x) - g(x)| \ dx = W_1(f,g) \numberthis \label{eq: MeanWD}
%\end{align*}
%From (\ref{eq: MeanWD}) we see that the addition of mean zero Gaussian noise does not on average affect our estimate of $W_1(\textbf{f}, \textbf{g})$. It is not possible to show, however, the same result when $ j \geq 2$, since for a non-linear function $g$, $\mathbb{E}(g(x)) \neq g(\mathbb{E}(x))$ in general. We now derive an upper bound on the variance of $W_1(\textbf{f},\textbf{g})$.
%\begin{align*}
%\text{var}(W_1(\textbf{f},\textbf{g})) &= \mathbb{E}\Big((W_1(\textbf{f},\textbf{g}) - W_1(f,g))^2 \Big)\\
%&= \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f} - \textbf{g}| - |f - g| \ dx \times \int_{-\infty}^\infty |\textbf{f} - \textbf{g}| - |f - g| dx \Big)\\
%&\leq \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f} - f| + |f - g| - |g-\textbf{g}| - |f-g| \ dx \times \int_{-\infty}^\infty |\textbf{f} - f| + |f - g| - |g-\textbf{g}| - |f-g| \ dx \Big)\\
%&=  \mathbb{E}\Big(\int_{-\infty}^\infty 2|v| \ dx \times \int_{-\infty}^\infty 2|v| \ dx \Big ) = 4. \numberthis
%\end{align*}

\noindent The WD is used to quantify the distance between  discrete and continuous functions, as shown in Figure (\ref{fig:1DWDCalculations}). In Figure \ref{fig: 1Ddiscreet} the WD between three points in computed, and in Figure \ref{fig: 1DContinuous} the WD between two Gaussians is computed.

% If $f$ and $g$ are the respective PDFs of
%\begin{equation}
%x \sim \mathcal{N}(\mu_x, \sigma^2_x) \text{   and   } y \sim \mathcal{N}(\mu_y, \sigma^2_y), 
%\end{equation} then \cite{Downson} proves that
%\begin{equation}
%W_2(f, g) = \sqrt{(\mu_y - \mu_x)^2 + (\sigma_y - \sigma_x)^2 }
%\label{eq: 1DGaussCalc}
%\end{equation}
%In Figure \ref{fig: 1DContinuous} we approximate (\ref{eq: 1DWD}) with $j = 2$ using the Trapezoidal method and obtain an error between (\ref{eq: 1DWD}) and (\ref{eq: 1DGaussCalc}) of $4.8246 \times 10^{-5}$.

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale = .49]{1st1DWD.pdf}
  \caption{Discrete Points, WD $\approx$ 1.7475}
  \label{fig: 1Ddiscreet}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale = .5]{2nd1DWD.pdf}
  \caption{Continuous Functions, WD $\approx$ 2.236}
  \label{fig: 1DContinuous}
\end{subfigure}
\caption{Computations of $W_2$ in one dimension}
\label{fig:1DWDCalculations}
\end{figure}

\section{Two Dimensional Case}
\label{sec: TwoD}

In two or more dimensions, there is no closed form expression for the Wasserstein distance. One prominent exception is in the case of two Gaussians. If 
\begin{equation}
x \sim \mathcal{N}(\mu_x, \Sigma_x) \text{   and   } y \sim \mathcal{N}(\mu_y, \Sigma_y), 
\end{equation}
and $f, g$ denote the PDFs of $x, y$ respectively, then \cite{Downson}  shows 
\begin{equation}
W_2(f, g) = |\mu_x - \mu_y| +\text{tr}(\Sigma_x + \Sigma_y - 2\sqrt{\Sigma_x\Sigma_y}).
\end{equation}

\noindent In this section a method to approximate the Wasserstein distance between discrete or continuous distributions in $\mathbb{R}^2$ is presented. Consider a physical domain with points $x_{i}, 1 \leq i \leq n \times m$, where $n, m$ are the number of points in the $x, y$ dimensions respectively. Form the matrix $X$ such that $(X)_{j, k}$ is the amount transported from $x_j$ under $f$ to $x_k$ under $g$. Form the matrix $(C)_{j, k} = |x_j - x_k|$ as the distance between points $x_j$ and $x_k.$  With this framework, computing the WD is then equivalent to solving 
\begin{equation}
\begin{aligned}
& {\text{minimize}}
& \sum_{ij}X_{ij}C_{ij} \\
& \text{subject to}
& X\textbf{1} = (a_1, \dotsc, a_m)^{T} \\ 
& & \textbf{1}^{T}X = (b_1, \dotsc, b_n)\\
& & (X)_{ij} \geq 0,
\end{aligned}
\label{eq: dual}
\end{equation}
where \textbf{1} indicates a vector of ones of appropriate length. Because of the linearity of the objective function, (\ref{eq: dual}) can be solved with the simplex method. In Figures \ref{fig: 2Ddiscreet} and \ref{fig: 2DContinuous}, the WD between discreet functions and continuous functions, respectively, in two dimensions is computed.

 \begin{figure}[H]
\centering
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[scale = .4]{2DPoints.jpg}
  \caption{Discrete Points, WD $\approx$ 5.69}
  \label{fig: 2Ddiscreet}
\end{subfigure}%
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[scale = .4]{2DGaussians.jpg}
  \caption{Continuous Functions, WD $\approx$ 7.12}
  \label{fig: 2DContinuous}
\end{subfigure}
\caption{Computations of WD in two dimensions}
\label{fig:1DWDCalculations}
\end{figure}
\noindent \textbf{Remark:} Although the continuous examples given were normal distributions, the WD can be computed between any two continuous functions.

\section{Application to Biological Systems}
\label{sec: BioSystems}
This section applies the tools from previous sections to quantify the differences between biological models that use particles to study larger phenomenon (``micro" models), such as the growth of cancer, and those that study such phenomenon on a macroscopic scope. This question is of critical importance to researchers, since the micro-level models are computationally cheaper and thus providing measures of their accuracy can be very useful.

\subsection{One Dimensional Case}
For illustration, we begin by examining the growth of cancer cells. To model such a phenomenon on a macroscopic level, we study the model
\begin{equation}
\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u},
\label{eq: MacroModel}
\end{equation}
where $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. The solution to ($\ref{eq: MacroModel}$) is the well-known Gaussian and is given by

\begin{equation*}
Macro_\rho(x) = \frac{1}{\sigma \sqrt{2\pi}}\exp\Big(\frac{-(x-\mu)^2}{2\sigma^2}\Big)
\end{equation*}

\noindent For our micro-model, we consider the model given by
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel}
\end{equation}
where $B_t$ denotes a standard Brownian motion, with $B(0) = 0$. In order to compare our micro-model to the macro-model, we approximate (\ref{eq: MicroModel}). For $\Delta t > 0$, we have
\begin{equation}
\int_0^{\Delta t} \ dX_t = \sigma \int_{0}^{\Delta t} \ dB_t \implies x_{\Delta t} - x_0 = \sigma \textbf{Z}
\end{equation}
where $\textbf{Z} \sim \mathcal{N}(0, \Delta t)$. Using properties of variance, 
\begin{equation}
x_{\Delta t} = x_0 + \sigma \sqrt{\Delta t}\textbf{W}
\label{eq: BrownianEQ}
\end{equation} 
where $\textbf{W} \sim \mathcal{N}(0, 1)$. To compute $Micro_{\rho}$, $N \in \mathbb{Z}^+$ i.i.d. realizations of (\ref{eq: BrownianEQ}) are computed, from which a histogram ($Micro_{\rho}$) is created over the physical domain.  In Figures \ref{fig: rhoInitial} and \ref{fig: rhoFinal} we plot $Macro_{\rho}, Micro_{\rho}$ at the initial and final times, respectively.
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{InitialDistributions.jpg}
  \caption{Solutions at initial time}
  \label{fig: rhoInitial}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{FinalDistribution.jpg}
  \caption{Solutions at final time}
  \label{fig: rhoFinal}
\end{subfigure}
\caption{$N = 10^3, \ T_{final} = 1, \ \Delta t =.01.$}
\label{fig:test}
\end{figure}

\noindent Using a standard normal PDF as the initial condition for $Macro_{\rho}$, and $\Delta x = .05$, we see in Figures \ref{fig:FittingError} and \ref{fig: FittingLogError} the decay of WD and log(WD) against number of particles, respectively.  All WD reported are at $T_{final} = 1, \ \Delta t = .01.$

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{WDVSN1D.jpg}
  \caption{Fit WD and N}
  \label{fig:FittingError}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{LogWDVSN1D.jpg}
  \caption{Fit Log(WD) and Log(N)}
  \label{fig: FittingLogError}
\end{subfigure}
\caption{Decay of WD with N}
\label{fig:test}
\end{figure}

\noindent We now explore how changing $\Delta x$, for a fixed $N$, changes the WD.

\begin{figure}[H]
\begin{center}
\includegraphics[scale = .4]{WDVSDX.jpg}
\end{center}
\caption{WD vs $\Delta x$.}
\end{figure}

\noindent We conclude from Figure \ref{fig:FittingError}  that the error between Macro$_\rho$ and Micro$_\rho$ follows the relationship 
\begin{equation*}
\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
\end{equation*}
 where $C \approx 3.33, \ \alpha \approx 0.44 $. When computing the linear fit  $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$. We thus conjecture that the Wasserstein distance between Micro$_\rho$ and Macro$_\rho$ decays at a rate of $1/\sqrt{N}$.
\begin{theorem}
$\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
\end{theorem}
\begin{proof}
Denote Macro$_\rho$ by $\rho_{mi}$, Micro$_\rho$ by $\rho_{ma}$. Then, for some $\phi \in C_b(\mathbb{R})$
\begin{equation*}
\langle \rho_{mi}, \phi \rangle = \int_{\mathbb{R}} \rho_{mi} \phi(x) \ dx = \frac{1}{N}\sum_x \phi(x_i(t)).
\end{equation*}
By the Law of Large Numbers, 
\begin{equation*}
\frac{1}{N}\sum_x \phi(x_i(t)) \xrightarrow{N \rightarrow \infty} \int_{\mathbb{R}} \phi(x)\rho_{mi}(x) \ dx + \frac{1}{\sqrt{N}}\text{std}(\phi(x))
\end{equation*}
where $\text{std}(\phi(x))$ denotes the standard deviation of $\phi$.
\end{proof}

\noindent We now consider a model with a source term, meaning the macro-level model is given by

\begin{equation}
\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u} + \rho(1-\rho),
\label{eq: MacroModel2}
\end{equation}

To calculate $\rho$Micro, $N$ realizations of (\ref{eq: BrownianEQ}) are computed. At each time-step, a new particle is created with probability $\Delta t(1-\rho)$ and destroyed with probability . A histogram ($Micro_{\rho}$) is then 
computed over the physical domain.

\subsection{Two Dimensional Case}

In two dimensions, the equation for the macro model becomes 
\begin{equation}
\frac{\partial \rho}{\partial t} = D\Big(\frac{\partial^2 \rho}{\partial^2 x} + \frac{\partial^2 \rho}{\partial^2 y}\Big) + \rho(1-\rho),
\label{eq: Macro2D}
\end{equation}
where $\rho(t,x,y)$ denotes the density of cancer cell at time $t$ at location $(x,y).$ Similarly, (\ref{eq: BrownianEQ}) becomes 
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel2}
\end{equation}
where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.

\section{Conclusion}
\label{sec: Conclusion}

\section{Definition: More for My Use}
\begin{itemize}
\item Let $X$ be a set. An algebra on $X$ is a collection $\Sigma \subseteq 2^X$ that is closed under finite set operations (complement, union, and intersection). A 
 $\sigma-$algebra on $X$ is a collection $\Sigma \subseteq 2^X$ that is closed under countably many such operations. The intersection of a collection of $\sigma-$algebras is a $\sigma-$algebra. 
 \item A \textbf{Borel set} in a topological space $(X, \mathscr{T})$ is a set that can be made from open sets $U \in \mathscr{T}$ by countable union, countable intersection, or relative complement. The collection of all Borel sets on $X$ forms a $\sigma-$algebra, known as the Borel $\sigma-$algebra, and is denoted by $\mathcal{B}(X)$.
 \item Let $X, Y$ be two topological spaces. A function $f: X \rightarrow Y$ is called a Borel map if 1) $f^{-1}(A)$, A open, is a Borel subset, 2) $f^{-1}(B)$, B closed, is a Borel subset, and 3) $f^{-1}(C)$, C Borel, is a Borel subset.
 \item Given a measurable space (X, $\Sigma$) and a measure $\mu$ on that space, a set $A \subset X$ in $\Sigma$ is called an \textbf{atom} if $\mu(A) > 0$ and for any measurable subset $B \subset A$ with $\mu(B) < \mu(A)$, the set B has measure zero. A measure that has no atoms is called \textbf{non-atomic}.
\end{itemize}

Informal sources:
\begin{itemize}
\item (1) http://www.math.umd.edu/~yanir/OT/AmbrosioGigliDec2011.pdf
\item (2) http://www.math.tohoku.ac.jp/~aida/workshop/H23-Daisympo/kuwada.pdf
\end{itemize}
\begin{thebibliography}{9}% maximum number of references (for label width)
\bibitem{CVillani}
Villani, C., Optimal transport, old and new, June 13, 2008, http://cedricvillani.org/wp-content/uploads/2012/08/preprint-1.pdf

\bibitem {AN}
Ambrosio, L., Gigli, N. A user's guide to optimal transport.  \url{http://www.math.umd.edu/~yanir/OT/AmbrosioGigliDec2011.pdf}.

\bibitem{Downson}
Dowson, D.C., B.V Landau, The Frechet Distance between Multivariate Normal Distributions, J. Multivariate Analysis 12, 450-455 (1982).

\bibitem{P}  
Prokhorov, Yu. V., Convergence of random processes and limit theorems in probability theory. (Russian) Teor. Veroyatnost. i Primenen. 1 (1956), 177-238.

\end{thebibliography}

\end{document}

% - Release $Name:  $ -

### 2016-05-23 model_1D/KPP_Micro.m

@@ -1,4 +1,4 @@
-%   function [rho,X] = KPP_Micro(rhoIC,x,D,dt,T,N,seedNum)
+   function [rho,X] = KPP_Micro(rhoIC,x,D,dt,T,N,seedNum)
 % 
 % Solve the KPP equation as a particle system:
 %    dX_t = σdB_t
@@ -42,7 +42,7 @@
         indexDivision = logical( (oneMrho>0).*(coin_M<dt*abs(oneMrho)) );
         %% 3) update
         if (length(indexDivision)>0)
-            X_new = X(indexDivision,:,:);
+            X_new = X(indexDivision,:);
         end
         if (length(indexKill)>0)
             X(indexKill,:) = [];
@@ -54,5 +54,5 @@
     end
     %-------------------------------------------%
     %-------------------------------------------%
-%  end
+ end
 

### 2016-05-23 model_1D/OneDStochasticFinal.m

@@ -5,16 +5,20 @@
 
 clc; clear; close all;
 
+% DeltaX = [2 1 .5 .25 .0125];
+% BigT = [10 10^2 10^3 10^4 10^5]
+% for m = 1:length(DeltaX)
+
 %% Parameter model
 D  = 1;                                % noive level
 sigma = sqrt(2*D);
 l0  = .2;                                % radius Initial condition
-num = [10^4];                          % number particles
+num = 10^3;                          % number particles
 a   = -5;                              % lower bound on grid
 b   = 5;                               % upper bound on grid
-dx1 =.05;                                % meshsize for micro
+dx1 = .25;                                % meshsize for micro
 dx3 = .1;                             % meshsize for Hist for rhoMicro
-dx2 =.05;                                % meshsize for macro
+dx2 =.25;                                % meshsize for macro
 intX = a:dx1:b;                        % physical discretization
 dt = .01;                              % delta t
 T =  0:dt:1;                           % time vector
@@ -45,13 +49,14 @@ for i = 1:length(num)
               WD_discreet(j)= WD_discreet_cont(X,intX,rhoMacro(:,j));
 %             [WDHist(j)] = WDFinal(HistCDF,MacroCDF,intXHist,intXMacro);
 %             [WDParticle(j),pCombined,FINV,GINV] = WDFinal(StepCDF,MacroCDF,FINALX{j},intXMacro);
-%            [Vel1(i,j), Vel2(i,j),xLoc1(j),xLoc2(j)] = CalculateVelocity(T,rhoMacro(:,j),rhoMicro(:,j),intX,xLoc1(j-1),xLoc2(j-1));
+%              [Vel1(i,j), Vel2(i,j),xLoc1(j),xLoc2(j)] = CalculateVelocity(T,rhoMacro(:,j),rhoMicro(:,j),intX,xLoc1(j-1),xLoc2(j-1));
 %             plot(abs(FINV(:,j)-GINV(:,j))); grid on; hold on;
       end
      toc
      hold off;
 end
-toc
+% toc
+
 
 % figure; plot(Vel1(2:end),'k--'); grid on; 
[…]

### 2016-05-23 model_2D/ComputeVelocity.m

@@ -1,0 +1,7 @@
+function [distance,xval,yval] = ComputeVelocity(x,y,rhoMacro,c)
+
+[Y,I] = min(abs(rhoMacro(:)-c));
+[row, col] = ind2sub(size(rhoMacro),I);
+distance = sqrt(x(row)^2 + y(col)^2);
+
+end

### 2016-05-23 model_2D/Micro.m

@@ -1,4 +1,4 @@
-%  function rho = Micro(x,y, xInit, yInit,D,dt,T,N)
+%   function rho = Micro(x,y, xInit, yInit,D,dt,T,N)
 % 
 % Solve the KPP equation as a particle system:
 %    dX_t = σdB_t

### 2016-05-23 model_2D/TwoDStochastic.m

@@ -1,4 +1,4 @@
-function TwoDStochastic
+% function TwoDStochastic
 % This function computes rhoMicro, rhoMacro for a given initial condition
 % in two dimensions. It calls Micro.m and Macro.m. 
 
@@ -7,33 +7,46 @@ clc; clear all; close all;
 
     dx = .2;
     dy = .2;
-    dt = .1;
-    T = 1;
+    dt = .01;
+    T = 2;
+    t = 0:dt:T;
     x = -4:dx:3;
     y = -4:dy:4;
-    D = 1;
+    D = 50;
     N = 1000;
+    c = .0005; % threshold for computing contours.
     
-    
-    mu1 = 1;
+    % Initial Conditiion for rhoMacro
+    mu1 = 0;
     mu2 = 0;
-    var1 = 1;
+    var1 = 2;
     var2 = 2;
     
-    xInit = normpdf(x,mu1,var1);
-    yInit = normpdf(y,mu2,var2);
-    
     % Call Micro, Macro
-    rhoMicro = Micro(x,y,xInit,yInit,D,dt,T,N); 
+%     rhoMicro = Micro(x,y,xInit,yInit,D,dt,T,N); 
     rhoMacro = Macro(x,y,D,dt,T,mu1,mu2,var1,var2);
-
-    subplot(2,2,1); surf(x,y,rhoMicro(:,:,1)'); colorbar;
-    xlabel('x'); ylabel('y'); title('Micro, Initial');
-    subplot(2,2,2); surf(x,y,rhoMacro(:,:,1)'); colorbar;
-    xlabel('x'); ylabel('y'); title('Macro, Initial')
-    subplot(2,2,3); surf(x,y,rhoMicro(:,:,end)');colorbar;
-    xlabel('x'); ylabel('y'); title('Micro, End')
-    subplot(2,2,4); surf(x,y,rhoMacro(:,:,end)');colorbar;
-    xlabel('x'); ylabel('y'); title('Macro, End')
     
-end
+    % Compute Velocities
+    for j = 1:floor(T/dt + .5)
+        distanceMacro(j) = ComputeVelocity(x,y,rhoMacro(:,:,j),c);
+    end
+    
+    Fit = polyfit(1:size(rhoMacro,3),distanceMacro./dt,1);
+    figure; plot(Fit);
+    xlabel('Time'); ylabel('Velocity');
+  
+    
+ 
+
+
+
[…]

### 2016-05-23 tests/MinimizeTableau.m

@@ -1,0 +1,71 @@
+%Author: Sebastien Motsch (smotsch@asu.edu).
+
+%This code takes takes a tableau and reduces it by Gaussian eleminitation
+%until it arrives at an optimal solution.
+
+function [A_up,b_up,X,bv]=MinimizeTableau(A,b,r,InitialCost,bvInit)
+% Solve the minimization problem:
+%     argmin C⋅X
+%       with AX = b, X≥0
+% The algorithm needs to start from a basic solution:
+%       A = ['Id' | A_b^-1⋅A_n]
+% Input:
+%     . A : m×n matrix, m constraints, n free variables
+%     . b : m vector, value of the constraint
+%     . r : gradient of the cost in the non-basic variable
+%             r = ∇_x_N (C⋅X) =c_n -  (A_b^-1 ⋅ A_n)^T c_b
+% Output:
+%     . Tableau: update tableau after row manipulation
+%     . X  : optimal solution (argmin)
+%     . bv : indices of the basic solutions
+%     
+    
+% init
+    [m n]= size(A);
+    Tableau = [A b;
+               r -InitialCost];
+     
+    if exist('bvInit', 'var')
+        bv = bvInit;
+    else
+        bv = zeros(m,1);
+    end
+    %-------------------------%
+    %   loop  minimization    %
+    %-------------------------%
+    tic
+    for iloop = 1:n
+        [valC j0] = min(Tableau(m+1,1:n)); % j0=ebv 'entering basic variable'
+        if valC < -1d-10
+            listRatio = Tableau(1:m,end)./Tableau(1:m,j0);
+            listRatio = listRatio + (Tableau(1:m,j0)<=0)*1d10;
+            [valRatio i0] = min(listRatio);                  % i0 lbv 'leaving basic variable'
+            %% update basic variables
+            bv(i0) = j0;
+        else
+            %fprintf('max!\n')
+            break;
+        end
+        Tableau(i0,:) = Tableau(i0,:)/Tableau(i0,j0);     % Tableau(i0,j0) = 1
+        for i = 1:(m+1)
+            if (i ~= i0)
+                Tableau(i,:) = Tableau(i,:) - Tableau(i,j0)*Tableau(i0,:);
+            end
+        end
+        %Tableau(end,:);
+    end
+    timeAlgo=toc;
+    %-------------------------%
+    %-------------------------%
+    % final
+
[…]

### 2016-05-23 tests/Simplex_OT.m

@@ -42,8 +42,8 @@ switch response
 
     case '3D'
 
-        Xpos = randn(4,3); % Starting points
-        Ypos = randn(4,3) + ones(4,1)*[5 0 1]; % Ending points
+        Xpos = randn(5,3); % Starting points
+        Ypos = randn(5,3) + ones(5,1)*[5 0 1]; % Ending points
         nX = length(Xpos(:,1));
         nY = length(Ypos(:,1));
              

### 2016-05-23 tests/TwoDGaussians.m

@@ -31,12 +31,12 @@
     clear all; 
     clc; close all;
 
-    mu_f    = [0 0];
+    mu_f    = [-2 -3];
     Sigma_f = [1 .3; .3 1];
-    mu_g    = [2 2];
+    mu_g    = [2 3];
     Sigma_g = [.25 .3; .3 1];
     
-    [X1,X2] = meshgrid(linspace(-5,5,7)',linspace(-5,5,7)');
+    [X1,X2] = meshgrid(linspace(-5,5,10)',linspace(-5,5,10)');
     XMESH = [X1(:) X2(:)];
     f = mvnpdf(XMESH,mu_f,Sigma_f);
     g = mvnpdf(XMESH,mu_g,Sigma_g);
@@ -45,8 +45,8 @@
     f = f/FSUM; g = g/GSUM;
     
     figure; 
-    surf(X1,X2,reshape(f,7,7)); hold on
-    surf(X1,X2,reshape(g,7,7)); hold off;
+    surf(X1,X2,reshape(f,10,10)); hold on
+    surf(X1,X2,reshape(g,10,10)); hold off;
     
     Xpos = XMESH;
     Ypos = Xpos;

### 2016-05-24 Paper/Paper.tex
updates

@@
 \usepackage{subfig}
+\usepackage{mwe}
 
@@
 \begin{equation}
-\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u},
+\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 x} + \rho(1-\rho),
 \label{eq: MacroModel}
 \end{equation}
-where $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. The solution to ($\ref{eq: MacroModel}$) is the well-known Gaussian and is given by
-
-\begin{equation*}
-Macro_\rho(x) = \frac{1}{\sigma \sqrt{2\pi}}\exp\Big(\frac{-(x-\mu)^2}{2\sigma^2}\Big)
-\end{equation*}
-
-\noindent For our micro-model, we consider the model given by
+where $\sigma$ is a scalar and $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. For the micro-model, we consider the equation
 \begin{equation}
@@
 \end{equation}
-where $B_t$ denotes a standard Brownian motion, with $B(0) = 0$. In order to compare our micro-model to the macro-model, we approximate (\ref{eq: MicroModel}). For $\Delta t > 0$, we have
+where $B_t$ denotes a standard Brownian process, with $B(0) = 0$. For $\Delta t > 0$, we have
 \begin{equation}
@@
 \noindent Using a standard normal PDF as the initial condition for $Macro_{\rho}$, and $\Delta x = .05$, we see in Figures \ref{fig:FittingError} and \ref{fig: FittingLogError} the decay of WD and log(WD) against number of particles, respectively.
+Figure \ref{fig: WDVsX} shows that WD decays linearly a $\Delta x$ decreases.
 All WD reported are at $T_{final} = 1, \ \Delta t = .01.$
 
-\begin{figure}[H]
+\begin{figure}
+\begin{subfigure}{.5\linewidth}
 \centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{WDVSN1D.jpg}
-  \caption{Fit WD and N}
+\includegraphics[scale = .4]{WDVSN1D.jpg}
+\caption{Fitting WD to N}
 \label{fig:FittingError}
 \end{subfigure}%
-\begin{subfigure}{.5\textwidth}
+\begin{subfigure}{.5\linewidth}
 \centering
-  \includegraphics[width=.9\linewidth]{LogWDVSN1D.jpg}
-  \caption{Fit Log(WD) and Log(N)}
+\includegraphics[scale = .4]{LogWDVSN1D.jpg}
+\caption{Fitting Log(WD) to Log(N)}
 \label{fig: FittingLogError}
+\end{subfigure}\\[1ex]
+\begin{subfigure}{\linewidth}
+\centering
+\includegraphics[scale = .4]{WDVSDX.jpg}
+\caption{Fitting WD to $\Delta x$}
+\label{fig: WDVsX}
 \end{subfigure}
-\caption{Decay of WD with N}
+\caption{One Dimensional WD Computations}
 \label{fig:test}
-\end{figure}
-
-\noindent We now explore how changing $\Delta x$, for a fixed $N$, changes the WD.
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[scale = .4]{WDVSDX.jpg}
-\end{center}
-\caption{WD vs $\Delta x$.}
 \end{figure}
@@
 \begin{equation*}
-\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
+\Big|W_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
 \end{equation*}
  where $C \approx 3.33, \ \alpha \approx 0.44 $.
-When computing the linear fit  $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$.
+When computing the fit $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$.
 We thus conjecture that the Wasserstein distance between Micro$_\rho$ and Macro$_\rho$ decays at a rate of $1/\sqrt{N}$.
 \begin{theorem}
-$\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
+$\Big|W_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
 \end{theorem}
@@
 
-\noindent We now consider a model with a source term, meaning the macro-level model is given by
+Figures \ref{fig: MicroVel} and \ref{fig: MacroVel} plot the respective velocities of $Micro_{\rho}, Macro_{\rho}$.
+To find the velocity between functions $f_1, f_2$ at time $t_j$, we set a value of $c = .1$, then find $x_{j+1} = f_{2}^{-1}(c),  x_{j} = f_{1}^{-1}(c)$.
+Then, the velocity at time $j$ is $v(j) = \frac{x_{j+1}- x_{j}}{\Delta t}$.
+Clearly, the velocity of $Micro_{\rho} \approx 1.38 \times 10^2$ is a random variable, while the velocity of $Macro_{\rho}\approx 1.5 \times 10^2$ is deterministic.
 
-\begin{equation}
-\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u} + \rho(1-\rho),
-\label{eq: MacroModel2}
-\end{equation}
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{1DMicroVelocity.jpg}
+  \caption{Fitting Micro$_{\rho}$ velocity}
+  \label{fig: MicroVel}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{1DMacroVelocity.jpg}
+  \caption{Fitting Macro$_{\rho}$ velocity}
+  \label{fig: MacroVel}
+\end{subfigure}
+\caption{Velocities, $T_{final} = 1, \Delta t = .01, N = 10^4, \sigma = \sqrt{2}.$}
+\label{fig:test}
+\end{figure}
 
-To calculate $\rho$Micro, $N$ realizations of (\ref{eq: BrownianEQ}) are computed.
-At each time-step, a new particle is created with probability $\Delta t(1-\rho)$ and destroyed with probability .
-A histogram ($Micro_{\rho}$) is then
-computed over the physical domain.
+We now investigate how the ratio $\frac{N}{\Delta x}$ impacts the WD.
+We consider the cases when $(N, \ \Delta x) = (10, 1), \ (10^2, .5), \ (10^3, .125), \ (10^4, .0625).$ For each case, $1,000$ i.i.id realizations of (\ref{eq: MicroModel}) are run.
+The WD between $Macro_{\rho}$ and $Micro_{\rho}$ at $T_{final = 1},\ \Delta t = .01$ is computed, and then the max of these WD values is computed.
+We plot the corresponding max WD and pairings of $(N, \ \Delta x)$ in Figures \ref{fig: FittingWDNDX} and \ref{fig: FittingLogWDNDX}.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{WDVSNDX.jpg}
+  \caption{Fitting WD to $\frac{N}{\Delta x}$}
+  \label{fig: FittingWDNDX}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{LogWDVSNDX.jpg}
+  \caption{Fitting Log(WD) to $Log(\frac{N}{\Delta x})$}
+  \label{fig: FittingLogWDNDX}
+\end{subfigure}
+\caption{One Dimensional WD Computations}
+\label{fig:test}
+\end{figure}
 
@@
 \end{equation}
-where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.
+where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.  Figures \ref{fig:Fitting2DMicroVelocity} and \ref{fig: Fitting2DMacroVelocity}
+plot the respective velocities of $Micro_{\rho}, Macro_{\rho}$.
+Clearly, the velocity of $Micro_{\rho} \approx $ is a random variable, while the velocity of $Macro_{\rho}\approx 1.5 \times 10^2$ is deterministic.
+
+\begin{figure}[H]
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{WDVSNDX.jpg}
+  \caption{Fitting WD to $\frac{N}{\Delta x}$}
+  \label{fig:Fitting2DMicroVelocity}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{2DMacroVelocity.jpg}
+  \caption{Fitting Log(WD) to $Log(\frac{N}{\Delta x})$}
+  \label{fig: Fitting2DMacroVelocity}
+\end{subfigure}
+\caption{2D Velocities, $T_{final} = 1, \Delta t = .01, N = 10^3, \sigma = \sqrt{2}.$}
+\label{fig:test}
+\end{figure}
 

### 2016-05-24 model_1D/CalculateVelocity.m

@@ -1,0 +1,13 @@
+  function [xPos1] = CalculateVelocity(f,intX,xLoc1,xLoc2)
+% This code computes the velocity of a function.
+% fix a a value of Y as the "reference point". Author: Shane Lubold
+% shane.lubold@asu.edu
+
+yStar = .1;                            %threshold
+xM = find(intX==0);
+fTemp = f(xM:end);           %cut and use only positive x values
+[~, loc1] = min(abs(fTemp-yStar)); 
+intXTemp = intX(xM:end);
+xPos1 = intXTemp(loc1);
+end
+  

### 2016-05-24 model_1D/OneDStochasticFinal.m

@@ -9,16 +9,31 @@ clc; clear; close all;
 % BigT = [10 10^2 10^3 10^4 10^5]
 % for m = 1:length(DeltaX)
 
+
+num = [5 10 10^2 10^3 10^4]; 
+dx = [2 1 .5 .25 .125]; % number particles
+
+
+for i = 1:length(num)
+    a   = -5;                              % lower bound on grid
+b   = 5; 
+    dx1 = dx(i);
+    intX = a:dx1:b; 
+    IC = normpdf(intX,0,1) ;
+    D  = 1;  
+    dt = .01;
+    T =  0:dt:1;  
+    rhoMacro  = KPP_Macro(IC,intX,D,dt,T,2);
+    for k = 1:1000
 %% Parameter model
 D  = 1;                                % noive level
 sigma = sqrt(2*D);
 l0  = .2;                                % radius Initial condition
-num = 10^3;                          % number particles
 a   = -5;                              % lower bound on grid
 b   = 5;                               % upper bound on grid
-dx1 = .25;                                % meshsize for micro
+dx1 = dx(i);                                % meshsize for micro
 dx3 = .1;                             % meshsize for Hist for rhoMicro
-dx2 =.25;                                % meshsize for macro
+dx2 =dx1;                                % meshsize for macro
 intX = a:dx1:b;                        % physical discretization
 dt = .01;                              % delta t
 T =  0:dt:1;                           % time vector
@@ -26,36 +41,34 @@ IC = normpdf(intX,0,1) ;                               %  Initial condition
 
 %% Solve  Macro
 % rhoMacro1 = MacroFinal(intX,dx2,T,dt,sigma,a,b,l0);
-tic
-rhoMacro  = KPP_Macro(IC,intX,D,dt,T,2);
-toc
+
+
 %% Solve  Micro
-tic
+
 xLoc1(1) = 0; xLoc2(1) = 0;             % Initialize position vectors
 
-for i = 1:length(num)
+
     %[rhoMicro,FINALX] = MicroFinalFast(IC,intX,D,dt,T,2);
     [rhoMicro,X] = KPP_Micro(IC,intX,D,dt,T,num(i));
     tic
-      for j = 2:length(T)
+%       for j = 2:length(T)
[…]

### 2016-05-24 model_2D/Micro.m

@@ -1,4 +1,4 @@
-%   function rho = Micro(x,y, xInit, yInit,D,dt,T,N)
+   function rho = Micro(x,y,D,dt,T,N)
 % 
 % Solve the KPP equation as a particle system:
 %    dX_t = σdB_t
@@ -59,11 +59,11 @@
     
     end
     
-    figure;
-    for j = 1:nT
-        surf(rho(:,:,j)); grid on;
-        xlabel('x'); ylabel('y'); zlabel('f(x,y)');
-        pause(.1);
-    end
+%     figure;
+%     for j = 1:nT
+%         surf(rho(:,:,j)); grid on;
+%         xlabel('x'); ylabel('y'); zlabel('f(x,y)');
+%         pause(.1);
+%     end
    
-%  end
+ end

### 2016-05-24 model_2D/TwoDStochastic.m

@@ -8,13 +8,13 @@ clc; clear all; close all;
     dx = .2;
     dy = .2;
     dt = .01;
-    T = 2;
+    T = 1;
     t = 0:dt:T;
     x = -4:dx:3;
     y = -4:dy:4;
     D = 50;
     N = 1000;
-    c = .0005; % threshold for computing contours.
+    c = .005; % threshold for computing contours.
     
     % Initial Conditiion for rhoMacro
     mu1 = 0;
@@ -23,7 +23,7 @@ clc; clear all; close all;
     var2 = 2;
     
     % Call Micro, Macro
-%     rhoMicro = Micro(x,y,xInit,yInit,D,dt,T,N); 
+    rhoMicro = Micro(x,y,D,dt,T,N); 
     rhoMacro = Macro(x,y,D,dt,T,mu1,mu2,var1,var2);
     
     % Compute Velocities

### 2016-05-24 tests/Simplex_OT.m

@@ -70,6 +70,7 @@ for k=1:nY
         A(k+nX, k + (k2-1)*nY) = 1;
     end
 end
+
 % We have an extra constraint (last equality is automatically satisfied)
 A(end,:) = [];
 b(end)   = [];

### 2016-05-24 tests/TwoDGaussians.m

@@ -36,7 +36,7 @@
     mu_g    = [2 3];
     Sigma_g = [.25 .3; .3 1];
     
-    [X1,X2] = meshgrid(linspace(-5,5,10)',linspace(-5,5,10)');
+    [X1,X2] = meshgrid(linspace(-5,5,5)',linspace(-5,5,5)');
     XMESH = [X1(:) X2(:)];
     f = mvnpdf(XMESH,mu_f,Sigma_f);
     g = mvnpdf(XMESH,mu_g,Sigma_g);
@@ -45,8 +45,8 @@
     f = f/FSUM; g = g/GSUM;
     
     figure; 
-    surf(X1,X2,reshape(f,10,10)); hold on
-    surf(X1,X2,reshape(g,10,10)); hold off;
+    surf(X1,X2,reshape(f,5,5)); hold on
+    surf(X1,X2,reshape(g,5,5)); hold off;
     
     Xpos = XMESH;
     Ypos = Xpos;

### 2016-05-26 model_2D/Micro.m
New Paper Updates, Code Updates

@@ -10,26 +10,14 @@
 % Solve the KPP equation as a particle system:
 %    dX_t = σdB_t
 %  + 'birth process' with intensity 1-ρ.
-% 
 
-    clc; clear all; close all;
-    dx = .1;
-    dy = .1;
-    dt = .1;
-    x = -3:dx:3;
-    y = -3:dy:3;
-    N = 10;
-    T = 0:dt:50;
-    nT = floor(T(end)/dt + .5);
-    dt = .1;
-    D = 1;
     [xmesh,ymesh] = meshgrid(x,y);
-    rho = zeros(numel(x),numel(y),nT);
+    rho = zeros(numel(x),numel(y),floor(T/dt + .5));
     
     X = [rand(N,1), rand(N,1)];
     rho(:,:,1) = hist3(X,[numel(x),numel(y)]);
    
-    for j = 2:nT
+    for j = 2:floor(T/dt + .5)
         %  Move
         M = length(X);
         X = X + sqrt(2*D)*sqrt(dt)*randn(M,2); 

### 2016-05-26 model_2D/TwoDOptimizationCode.m

@@ -1,0 +1,84 @@
+function [final_cost, maxDiff1, maxDiff2] = TwoDptimizationCode(x,y,f,g)
+
+
+    [X1, X2] = meshgrid(x,y);
+    XMESH = [X1(:) X2(:)];
+    Xpos = XMESH;
+    Ypos = Xpos;
+    nX = length(Xpos);
+    nY = length(Ypos);
+
+    b = [f(:); g(:)];
+    
+    A = zeros(nX+nY,nX*nY);
+    for k=1:nX
+        A(k, (k-1)*nY+[1:nY]) = 1;
+    end
+    for k=1:nY
+        for k2=1:nX
+            A(k+nX, k + (k2-1)*nY) = 1;
+        end
+    end
+    
+%      % We have an extra constraint (last equality is automatically satisfied)
+      A(end,:) = [];
+      b(end)   = [];
+     
+    nX = length(Xpos(:,1));
+    nY = length(Ypos(:,1));
+    matrix_C = zeros(nX,nY);
+    for i=1:nX
+        for j=1:nY
+            matrix_C(i,j) = sqrt( (Ypos(j,1)-Xpos(i,1))^2 + (Ypos(j,2)-Xpos(i,2))^2 );
+        end
+    end
+    C = matrix_C(:);
+    
+    fprintf('---------------------------------------\n')
+    fprintf('Phase 1: find a basic feasible solution\n')
+    fprintf('---------------------------------------\n')
+    [A_phase2,b_phase2,X,bv] = MinimizeTableau(A,b,-sum(A),-sum(b));
+    r = C';
+    
+    for i=1:length(bv)
+        % find associated basic variable
+        j = bv(i);
+        if (j>0)
+            % update cost
+            r = r - C(j)*A_phase2(i,:);
+        end
+    end
+    
+    fprintf('---------------------------------------\n')
+    fprintf('Phase 2: find the optimal solution\n')
+    fprintf('---------------------------------------\n')
+    initial_cost = C'*X;
+    [A_opt,b_opt,X_opt,bv_opt] = MinimizeTableau(A_phase2,b_phase2,r,initial_cost,bv);
+    final_cost = C'*X_opt;
+    
+    
+    matrix_X_opt = reshape(X_opt,nY,nX)';
+    clf
+    hold on
+    plot(Xpos(:,1),Xpos(:,2),'bo',Ypos(:,1),Ypos(:,2),'rx')
+    for i=1:nX
[…]

### 2016-05-26 model_2D/TwoDStochastic.m

@@ -7,20 +7,22 @@ clc; clear all; close all;
 
     dx = .2;
     dy = .2;
-    dt = .01;
+    dt = .1;
     T = 1;
     t = 0:dt:T;
-    x = -4:dx:3;
-    y = -4:dy:4;
-    D = 50;
-    N = 1000;
+    x = -1:dx:1;
+    y = -1:dy:1;
+%     x= linspace(-5,5,7);
+%     y = x;
+    D = 1;
+    N = 10^2;
     c = .005; % threshold for computing contours.
     
     % Initial Conditiion for rhoMacro
-    mu1 = 0;
-    mu2 = 0;
-    var1 = 2;
-    var2 = 2;
+    mu1 = -1;
+    mu2 = 3;
+    var1 = 1;
+    var2 = 3;
     
     % Call Micro, Macro
     rhoMicro = Micro(x,y,D,dt,T,N); 
@@ -34,11 +36,12 @@ clc; clear all; close all;
     Fit = polyfit(1:size(rhoMacro,3),distanceMacro./dt,1);
     figure; plot(Fit);
     xlabel('Time'); ylabel('Velocity');
-  
     
- 
-
-
+    %Compute the WD
+    [WD,Diff1,Diff2] = TwoDOptimizationCode(x,y,rhoMicro(:,:,end),...
+    rhoMacro(:,:,end))
+    
+  
 
 %     subplot(2,2,1); surf(x,y,rhoMicro(:,:,1)'); colorbar;
 %     xlabel('x'); ylabel('y'); title('Micro, Initial');

### 2016-05-26 tests/TwoDGaussians.m

@@ -32,11 +32,11 @@
     clc; close all;
 
     mu_f    = [-2 -3];
-    Sigma_f = [1 .3; .3 1];
+    Sigma_f = [3 1; 1 1];
     mu_g    = [2 3];
     Sigma_g = [.25 .3; .3 1];
     
-    [X1,X2] = meshgrid(linspace(-5,5,5)',linspace(-5,5,5)');
+    [X1,X2] = meshgrid(linspace(-5,5,7)',linspace(-5,5,7)');
     XMESH = [X1(:) X2(:)];
     f = mvnpdf(XMESH,mu_f,Sigma_f);
     g = mvnpdf(XMESH,mu_g,Sigma_g);
@@ -45,8 +45,8 @@
     f = f/FSUM; g = g/GSUM;
     
     figure; 
-    surf(X1,X2,reshape(f,5,5)); hold on
-    surf(X1,X2,reshape(g,5,5)); hold off;
+    surf(X1,X2,reshape(f,7,7)); hold on
+    surf(X1,X2,reshape(g,7,7)); hold off;
     
     Xpos = XMESH;
     Ypos = Xpos;

### 2016-06-03 Paper/Paper.tex
New Paper Updates, Code Updates
implemented explicit finite difference methods for Dirichlet and Neumann BC

@@
 
+\begin{center}
+\textbf{Recent progress:}
+\end{center}
+
+\begin{enumerate}
+\item Went through the 1D,2D non-bio work. All data files for images in 1D section have been saved to folder ``Data for Paper''. Also fixed wording in section. 
+\item For 1D bio work, saved data in folder ``Data for Paper''
+\begin{itemize}
+\item Put in images of rhoMicro, rhoMacro at time initial and final. 
+\item Fitted WD to N, Log(WD) to Log(N) and fitted WD to $\Delta x$, saved data.
+\item Computed rhoMicro, rhoMacro velocity, included images, saved data.
+\end{itemize}
+\item For 2D bio work, 
+\begin{itemize}
+\item Put in some basic wording
+\item Wrote Velocity code, included Macro velocity image and data in folder. 
+\item Fixed micro code.
+\item Added WD code for 2D (took old simplex method code and call it in 2DStochasticFinal).
+\end{itemize}
+\textbf{Questions:} 
+\begin{itemize}
+\item Normalization of Micro
+\item How large are our systems for the 2D bio? Even 121 variables takes 4.5 minutes to compute. Might try hungarian algorithm, interior point methods, etc to speed up.
+\end{itemize}
+\end{enumerate}
+
+\newpage
 \section{Introduction}
@@
 \section{One Dimensional Case} \label{sec: OneD}
-In one dimension, there exists a closed form solution to computing the Wasserstein distance between two probability density functions. Let $f, g: \mathbb{R} \rightarrow \mathbb{R}$ such that
+In one dimension, there exists a closed form solution to computing the Wasserstein distance between two probability density functions.
+Let $f, g: \mathbb{R} \rightarrow \mathbb{R}$ be functions such that
 \begin{equation}
-\int_{\mathbb{R}} f \ dx = \int_{\mathbb{R}} g \ dx.
+\int_{\mathbb{R}} f(x) \ dx = \int_{\mathbb{R}} g(x) \ dx.
 \end{equation}
@@
   \centering
-  \includegraphics[scale = .49]{1st1DWD.pdf}
+  \includegraphics[scale = .4]{1st1DWD.pdf}
   \caption{Discrete Points, WD $\approx$ 1.7475}
@@
   \centering
-  \includegraphics[scale = .5]{2nd1DWD.pdf}
-  \caption{Continuous Functions, WD $\approx$ 2.236}
+  \includegraphics[scale = .4]{1DWDNormal.jpg}
+  \caption{Continuous Functions, WD $\approx$ 2.06}
   \label{fig: 1DContinuous}
 \end{subfigure}
-\caption{Computations of $W_2$ in one dimension}
+\caption{Computations of WD in one dimension}
 \label{fig:1DWDCalculations}
@@
   \includegraphics[width=.9\linewidth]{WDVSNDX.jpg}
-  \caption{Fitting WD to $\frac{N}{\Delta x}$}
+  \caption{WRONG IMAGE HERE}
   \label{fig:Fitting2DMicroVelocity}
@@
 \end{document}
-
-% - Release $Name:  $ -

### 2016-06-03 model_2D/Macro.m

@@ -1,55 +1,127 @@
-function rhoFINAL = Macro(x,y,D,dt,T,mu1,mu2,var1,var2)
+% function rhoFINAL = Macro(x,y,D,dt,T,mu1,mu2,var1,var2)
 
-% This code solves the equation: du/dt = D(d^2u/dx^2 + d^2u/dy^2) + u(1-u) 
-% using a Crank-Nicolson scheme. Take dx = dy for simplicity. This code
+% This code solves the equation: du/dt = D(d^2u/dx^2 + d^2u/dy^2) + u(1-u)
+% with two approaches. The first uses finite differencing two solve under
+% both Dirichlet and Neumann BC. The second uses a Crank-Nicolson under 
+% an unknown BC. This code
 % calls blktridiag, (http://www.mathworks.com/matlabcentral/fileexchange/
 % 10603--block--tri-diagonal-matrices), which quickly constructs a
 % tridiagional matrix from submatrices. The resulting equation is
 % Au^{n+1} = Bu^{n} + f(u^{n}) so u^{n+1} = inv(A)*(Bu^{n}+dt*f(u))
 
 % Author: Shane Lubold (shane.lubold@asu.edu).
-
-    % Model
-    f = @(y) y.*(1-y);
-
+    clc; clear all; close all;
+    f = @(u) u.*(1-u);    
+    dx = 1.5;
+    dy = 1.5;
+    x = -5:dx:5;
+    y = -5:dy:5;
+    mu1 = 0;
+    mu2 = 0;
+    var1 = .2;
+    var2 = .2;
+    nX = length(x);
+    nY = length(y);
+    T = 20;
+    dt = 10^-2;
+    D = 1;
+    nT = floor(T/dt + .5);
     [X1,X2] = meshgrid(x,y);
     X1_tp = X1';
     X2_tp = X2';
-    rhoIC = mvnpdf([X1_tp(:) X2_tp(:)],[mu1, mu2],[var1, var2]);
+    rhoIC = reshape(mvnpdf([X1_tp(:) X2_tp(:)],[mu1, mu2],[var1, var2]),nX,nY);
+
+    % If we want Dirichlet BC (boundaries @ 0), use the following code
[…]

### 2016-08-16 model_1D/Micro_Macro.m
Examined Code

@@ -10,12 +10,12 @@
 D = 2/3;
 
 % Numerical parameters
-dt = .01;
-T  = 10;
-dx = .05;
-L  = 20;
+dt = .1;
+T  = 2;
+dx = .1;
+L  = 10;
 method_Kpp = 2;                           % 1: explicit, 2: Cranck-Nicholson
-Ninit  = 1e4;                           % number of particles initially
+Ninit  = 1e6;                           % number of particles initially
 seedNum = 3;
 % Initial condition
 x = -L:dx:L;
@@ -39,15 +39,15 @@ addpath('lib')
 nT = length(rhoMacro(1,:));
 stock_WD = zeros(1,nT);
 for k = 1:nT
-    stock_WD(k) = WD_Cont(rhoMicro(:,k)',rhoMacro(:,k)',x);
+    stock_WD(k) = WD_Cont(rhoMicro(:,k),rhoMacro(:,k),x);
 end
 %% B.2) Velocity traveling wave
 xStar_micro = zeros(1,nT);
 xStar_macro = zeros(1,nT);
 nX_half = floor(length(x)/2);
 for k = 1:nT
-    xStar_micro(k) = interp1(rhoMicro(nX_half:end,k)',x(nX_half:end),1/2,'left');
-    xStar_macro(k) = interp1(rhoMacro(nX_half:end,k)',x(nX_half:end),1/2,'left');
+    xStar_micro(k) = interp1(x(nX_half:end),rhoMicro(nX_half:end,k),1/2,'left');
+   xStar_macro(k) = interp1(x(nX_half:end),rhoMacro(nX_half:end,k),1/2,'left');
 end
 
 %---------------------------------------------------------%
@@ -61,6 +61,7 @@ plot(intT,stock_WD,'r','linewidth',2)
 xlabel('time')
 ylabel('WD')
 title('Wasserstein distance micro/macro over time')
+grid on;
 
 figure(2)
 if (shouldPlot)
@@ -69,23 +70,25 @@ else
     jumpPlot = nT-1;
 end
 for k = 1:jumpPlot:nT
-    plot(x,rhoMacro(:,k),'r','linewidth',2, ...
-         x,rhoMicro(:,k),'b','linewidth',2); 
+    plot(x,rhoMacro(:,k),'r','linewidth',2); hold on
+    plot(x,rhoMicro(:,k),'b','linewidth',2); 
+    hold off;
     legend('Macro','Micro'); 
[…]

### 2016-08-16 model_2D/lib/TwoDStochastic.m

@@ -12,13 +12,13 @@ N = 10^4;                               % number of cells (initially)
 Mass_rhoIC = 2;
 seedNum = 1;
 % domain
-dx = 1;
-dy = 1;
+dx = .8;
+dy = .8;
 x = -2:dx:2;
 y = -2:dy:2;
 % time
 dt = 10^-2;
-T =   5;
+T =   1;
 t = 0:dt:T;
 
 % Initial Conditiion
@@ -61,7 +61,8 @@ SimplexCode = toc;
  [xval, fval,exitflag] = linprog(C,[],[],A,b,zeros(size(C)),[]);
  LinProgCode = toc;
  DiffInMethods = abs(WD-fval)
-    
+   
+ xval = reshape(xval,length(x)^2, length(y)^2);
 
 %     subplot(2,2,1); surf(x,y,rhoMicro(:,:,1)'); colorbar;
 %     xlabel('x'); ylabel('y'); title('Micro, Initial');

### 2016-08-18 Paper/Manuscript/Shane's Part/ShaneManuscript.tex
Updated Data

--- 2016-05-23 Paper/Paper.tex
@@
 
-\newshadetheorem{thm}{Theorem}
-\definecolor{shadethmcolor}{HTML}{EDF8FF}
-\definecolor{shaderulecolor}{HTML}{45CFFF}
-\setlength{\shadeboxrule}{.4pt}
+\begin{document}
 
-% replace 'n' with the number of the homework assignment
-\title{Analyzing Biological Models with the Wasserstein Distance}
-
-% replace 'your name here' with your name
-\author{
-  Shane Lubold\\
-  \texttt{shane.lubold@asu.edu}\\
-  Arizona State University
-  \and
-  Sebastien Motsch \\
-  \texttt{smotsch@asu.edu}\\
-  Arizona State University
-}
-\date{\today}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-In recent years a number of mathematical models have been proposed that describe biological systems.
-One needs a way to compare the accuracy of these methods, in relation to each other and in relation to empirical data.
-The Wasserstein distance is a new metric that can be used for such a purpose.
-In one dimension, there exists a closed form expression for the Wasserstein distance, but in two or more dimensions, no such expression exists for most functions.
-In this paper we present a method for approximating the Wasserstein distance in two or more dimensions.
-We also consider several biological models and demonstrate how the Wasserstein distance can be used to analyze the effictiveness of such models.
-\end{abstract}
-
-\section{Introduction}
-
-  Modeling biological systems has increased in popularity in recent years.
-From bacteria to swarm bees, a number of models have been proposed to describe each system.
-But the increase in number of models does not necessary improve the comprehension of the underlying population.
-One needs to be able to compare there predictions and more important to measure their agreement with experimental data.
-One way to do this is to use the Wassertstein distance, which allows us to compare discrete and continuous dynamics (i.e.e, systems of ODEs and PDEs).
-But so far Wasserstein distance has been used mainly as a powerful tool to prove analytic results (i.e.
-existence/uniqueness of solutions).
-In this paper we use the Wasserstein distance to compare the accuracy of biological models in relation to empirical data.
-
-The challenge is that the Wasserstein distance, defined as a minimization problem,
-can only be computed explicity in one dimension.
-In two or higher dimensions, we must rely on an approximation.
-In this paper, we develop a numerical method that approximates the Wasserstein distance through the Simplex method for continuous distributions, and the Hungarian algorithm for discrete distributions (sets of points).
-Once the numerical methods have been implemented and tested, we will use the Wasserstein distance to perform data-model comparison for several dynamics (e.g.
-pedestrian dynamics, swarming model).
-In many studies, data-model comparisons are limited to only visual or qualitative agreement.
-This framework aims at providing quantitative measurement for the agreement (or discrepancy) between a model and experimental data.
-
-The rest of the paper is organized as follows.
-In Section \ref{sec: Preliminaries}, we present the general problem, outline the theory and notation, and provide some theoretical results regarding the Wasserstein Distance.
-In Section \ref{sec: OneD}, we provide the theory and computations of the Wasserstein distance in one dimension.
-We also consider the case of noisy data and show how this affects the computation of the Wasserstein distance.
-Section \ref{sec: TwoD} outlines the problem in two dimensions, and shows how the simplex method may be used to compute the Wasserstein distance between continuous and discrete distributions.
-In Section \ref{sec: BioSystems} we examine how models used to study biological systems compare to empirical data through the use of the Wasserstein distance.
-Section \ref{sec: Conclusion} provides concluding remarks and points to future work.
-
-\section{Literature Review} 
-The optimal transport problem, the problem of moving a mass from one location to another in the most efficient way possible, has been studied for several centuries.
-One way to phrase the problem is this.
-Consider a number $n$ of bakeries, each of which makes a certain amount of bread, and we know that there are $m$ restaurants that will consume a fixed amount of bread.
-Since the amount of bread produced and consumed is fixed for each bakery and restaurant, we can model these quantities as probability measures and create a ``density of production" and a ``density of consumption" on some space, which we can consider to be a metric space with the distance between a bakery and a restaurant as the length of the straight line between them.
-The optimal transport problem then seeks to move the bread from the bakeries to the restaurants by minimizing the total cost of transportation.
-For a more complete description of this problem, see Section 3 of \cite{CVillani}.
-
-\noindent Recent work has shown a number of theoretical properties of the Wasserstein distance, including the existence of solutions to the optimal transport problem under certain assumptions, \cite{CVillani}, the convexity of the solution space, \cite{AN}, and the weak convergence of measures, \cite{AN}.
-
-\noindent In recent years there has been a proliferation of biological models concerning a vast array of biological systems.
-
-\section{Preliminaries}
-\label{sec: Preliminaries}
-Given a Polish (complete, separable, metric) space $(M,d)$, define $\mathcal{P}(M)$ as the set of probability measures on $(M, \mathcal{B}(M))$.
-If $X, Y$ are two Polish spaces, $T$ is a Borel map, and $\phi \in \mathcal{P}(X)$ is a measure, we can define $T_{\#}\phi \in \mathcal{P}(Y)$, which we call the \textit{pushforward of $\phi$ through T}, as
-\begin{equation}
-T_{\#}\phi(A) = \phi(T^{-1}(A)), \ \ \ \ \ \forall A \subseteq Y, \  A \  \text{is Borel}.
-\label{eq: TransportDefn}
-\end{equation}
-The pushforward of $\phi$ through $T$ has the property that
-\begin{equation}
-\int_Y h \ dT_{\#}u = \int_X h \circ T d\mu,
-\label{eq: TransportProperty}
-\end{equation}
-for every Borel function $h: Y \rightarrow \mathbb{R} \cup \{\pm \infty\}$.
-Define a Borel cost function $c: X \times Y \rightarrow \mathbb{R} \cup \{\infty\}$.
-Following the notation of \cite{AN}, we call the following problem formulation the \textit{Monge optimal transport problem}.
-
-\noindent \textbf{Monge Optimal Transport Problem:} Let $v \in \mathcal{P}(X)$, $\mu \in \mathcal{P}(Y).$ We define the $j$th Wasserstein distance between $v$ and $\mu$ as
-\begin{equation}
-W_j(\mu,\nu) = \inf_{T} \sqrt[\leftroot{-2}\uproot{2}j]{\int_{M}c(x, T(x))^{j} \ d\mu(x)},
-\label{eq: CostWassDist}
-\end{equation}
-where the $\inf$ is taken over all $T$ satisfying (\ref{eq: TransportProperty}).
-As noted in \cite{AN}, Monge's definition can be ill-posed if no such admissible T exists (if $\mu$ is a Dirac delta and $\nu$ is not) or if the constraint that $T_{\#}\mu = \nu$ is not weakly sequentially closed, w.r.t any reasonable weak topology.
-To resolve these issues, we also present the formulation attributed to Kantorovich, \cite{AN}.
-
-\noindent \textbf{Kantorovich Optimal Transport Problem:}
-\noindent For each $\mu \in \mathcal{P}(X)$ and $\nu \in \mathcal{P}(Y)$, let $\Pi(f,g)$ denote the collection of couplings between $\mu$ and $\nu$. In other words, 
-\begin{equation}
-\Pi(f, g) = \Big\{\pi: \pi(A \times M) = \mu(A), \ \text{and} \ \pi(A \times M) = \nu(A), \ \text{for each $A \in \mathcal{B}(M)$}  \Big\}.
-\label{eq: Couplings}
-\end{equation}
-We then define the $j$-th Wasserstein distance between $\mu$ and $\nu$ as 
-\begin{equation}
-W_j(f, g) = \inf_{\pi \in \Pi(f, g)}\sqrt[\leftroot{-2}\uproot{2}j]{\int_{M \times M} d(x,y)^{j} d\pi(x, y)},
-\label{eq: CouplingsWassDist}
-\end{equation}
-where the $\inf$ is taken over all $\pi \in \Pi(f, g)$ that satisfy (\ref{eq: Couplings}).
-In some cases, $W_p(\mu, \nu) = \infty$, so we will restrict ourselves $W_j(f, g)$ to $\mathcal{P}_j(M) \subset \mathcal{P}(M)$, where $\mathcal{P}_j(M)$ denotes the set of probability measures on M with finite $j$-th moment.
-In other words,
-\begin{equation}
-\mathcal{P}_j(M) = \Big\{f \in \mathcal{P}(M): \int_M d(x, y)^j f(dy) < \infty, \ \text{for some $x \in M$} \Big\}.
-\label{eq: jPM}
-\end{equation}
-As shown in \cite{AN}, $\Pi(f, g)$ is a convex set that contains $f \times g$.
-If $c$ is lower semicontinuous and bounded below, then there exists a $\pi \in \Pi(f, g)$ that satisfies (\ref{eq: CouplingsWassDist}).
-\cite{AN} also proves that if $c$ is continuous and $\mu$ is non atomic, then the values of (\ref{eq: CostWassDist}) and (\ref{eq: CouplingsWassDist}) are equal.
-It is also possible to show that $W_{j}$ defines a distance function on $\mathcal{P}_j(M)$.
-Clearly $W_{j}(\mu, \nu) = W_{j}(\nu, \mu)$ and $W_{j}(\nu, \mu) \geq 0$, with equality only holding in the case that $\mu = \nu$.
-To prove that $W_{j}(\mu, \nu) \leq W_{j}(\nu, \psi) + W_{j}(\psi, \nu)$ for some measure $\psi$, it's convenient to use the so-called gluing lemma, \cite{CVillani}.
-
-\noindent \textbf{Gluing Lemma:} Let $(X_i, \mu_i), i = 1, 2, 3,$ be Polish probability spaces.
-If $(X_1, X_2)$ is a coupling of $(\mu_1, \mu_2)$ and $(Y_2, Y_3) $ is a coupling of $(\mu_2, \mu_3)$, then it is possible to construct a triple of random variables $(Z_1, Z_2, Z_3)$ such that $(Z_1, Z_2)$ has the same law as $(X_1, X_2)$ and $(Z_2, Z_3)$ has the same law as $(Y_2, Y_3)$.
-
-\noindent For the remainder of the paper, we restict ourselves to $(M,d) = (\mathbb{R}^n, d)$, with $n \geq 1$ and $d$ being the standard metric on $\mathbb{R}^n$.
-
-\section{One Dimensional Case} \label{sec: OneD}
-In one dimension, there exists a closed form solution to computing the Wasserstein distance between two probability density functions. Let $f, g: \mathbb{R} \rightarrow \mathbb{R}$ such that
-\begin{equation}
-\int_{\mathbb{R}} f \ dx = \int_{\mathbb{R}} g \ dx.
-\end{equation}
-If we define
-\begin{equation}
-F(x) = \int_{- \infty}^x f(s) \ ds \ \ \ \ \text{and} \ \ \ \ G(x) = \int_{- \infty}^x g(s) \ ds,
-\end{equation}
-then the $2$nd Wasserstein distance between $f$ and $g$ is
-\begin{equation}
-W_2^2(f, g) = {\int_{0}^1 |F^{-1}(y) - G^{-1}(y)|^2 \ dy}.
-\label{eq: 1DWD}
-\end{equation}
-From now on, we will refer to the value of ($\ref{eq: 1DWD}$) as the Wasserstein distance between $f$ and $g$, or simply the WD between $f$ and $g$.
-One can think of (\ref{eq: 1DWD}) as moving the $j$-th percentile of $f$ to the $j$-th percentile of $g$ (Figure \ref{fig: 1DWD}).
-
-\begin{figure}[H]
-\centering
-\includegraphics[scale = .4]{2ndPPercentile.pdf}
-\caption{Moving $j$-th percentile (shaded) of $f$ to $j$-th percentile of $g$.}
-\label{fig: 1DWD}
-\end{figure}
-%In some cases, it is possible the data one collects are corrupted by noise, and a relavent question is what are the mean and variance of the resulting approximation of the Wasserstein distance.
-Assume we observe $\textbf{f}$, $\textbf{g}$\footnote{From here on, boldface variables denote random variables.}, and we want to know the expected value and variance of our estimates of $W_1(f,g)$.
-%\begin{equation}
-%\textbf{f}(x) = f(x) + v, \ \ \ \textbf{g}(x) = g(x) + v, \ \ \ \ v \sim \mathcal{N}(0, \rho^2).
-%\end{equation}
-%When $j = 1$, (\ref{eq: Alt1DWD}) becomes
-%\begin{equation}
-%W_1(f, g) = \int_{-\infty}^\infty |f(x) - g(x)| \ dx.
-%\end{equation}
-%By linearity, we have that
-%\begin{align*}
-%\mathbb{E}(W_1(\tilde{f}, \tilde{g})) &= \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f}(x) - \textbf{g}(x)| \ dx \Big) \\  & = 
-%\int_{-\infty}^\infty |\mathbb{E}(\textbf{f}(x)) - \mathbb{E}(\textbf{g}(x))| \ dx \\
-%&= \int_{-\infty}^\infty |f(x) - g(x)| \ dx = W_1(f,g) \numberthis \label{eq: MeanWD}
-%\end{align*}
-%From (\ref{eq: MeanWD}) we see that the addition of mean zero Gaussian noise does not on average affect our estimate of $W_1(\textbf{f}, \textbf{g})$.
-It is not possible to show, however, the same result when $ j \geq 2$, since for a non-linear function $g$, $\mathbb{E}(g(x)) \neq g(\mathbb{E}(x))$ in general.
-We now derive an upper bound on the variance of $W_1(\textbf{f},\textbf{g})$.
-%\begin{align*}
-%\text{var}(W_1(\textbf{f},\textbf{g})) &= \mathbb{E}\Big((W_1(\textbf{f},\textbf{g}) - W_1(f,g))^2 \Big)\\
-%&= \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f} - \textbf{g}| - |f - g| \ dx \times \int_{-\infty}^\infty |\textbf{f} - \textbf{g}| - |f - g| dx \Big)\\
-%&\leq \mathbb{E}\Big(\int_{-\infty}^\infty |\textbf{f} - f| + |f - g| - |g-\textbf{g}| - |f-g| \ dx \times \int_{-\infty}^\infty |\textbf{f} - f| + |f - g| - |g-\textbf{g}| - |f-g| \ dx \Big)\\
-%&=  \mathbb{E}\Big(\int_{-\infty}^\infty 2|v| \ dx \times \int_{-\infty}^\infty 2|v| \ dx \Big ) = 4. \numberthis
-%\end{align*}
-
-\noindent The WD is used to quantify the distance between  discrete and continuous functions, as shown in Figure (\ref{fig:1DWDCalculations}).
-In Figure \ref{fig: 1Ddiscreet} the WD between three points in computed, and in Figure \ref{fig: 1DContinuous} the WD between two Gaussians is computed.
-
-% If $f$ and $g$ are the respective PDFs of
-%\begin{equation}
-%x \sim \mathcal{N}(\mu_x, \sigma^2_x) \text{   and   } y \sim \mathcal{N}(\mu_y, \sigma^2_y), 
-%\end{equation} then \cite{Downson} proves that
-%\begin{equation}
-%W_2(f, g) = \sqrt{(\mu_y - \mu_x)^2 + (\sigma_y - \sigma_x)^2 }
-%\label{eq: 1DGaussCalc}
-%\end{equation}
-%In Figure \ref{fig: 1DContinuous} we approximate (\ref{eq: 1DWD}) with $j = 2$ using the Trapezoidal method and obtain an error between (\ref{eq: 1DWD}) and (\ref{eq: 1DGaussCalc}) of $4.8246 \times 10^{-5}$.
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale = .49]{1st1DWD.pdf}
-  \caption{Discrete Points, WD $\approx$ 1.7475}
-  \label{fig: 1Ddiscreet}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale = .5]{2nd1DWD.pdf}
-  \caption{Continuous Functions, WD $\approx$ 2.236}
-  \label{fig: 1DContinuous}
-\end{subfigure}
-\caption{Computations of $W_2$ in one dimension}
-\label{fig:1DWDCalculations}
-\end{figure}
-
-\section{Two Dimensional Case}
-\label{sec: TwoD}
-
-In two or more dimensions, there is no closed form expression for the Wasserstein distance. One prominent exception is in the case of two Gaussians. If 
-\begin{equation}
-x \sim \mathcal{N}(\mu_x, \Sigma_x) \text{   and   } y \sim \mathcal{N}(\mu_y, \Sigma_y), 
-\end{equation}
-and $f, g$ denote the PDFs of $x, y$ respectively, then \cite{Downson}  shows 
-\begin{equation}
-W_2(f, g) = |\mu_x - \mu_y| +\text{tr}(\Sigma_x + \Sigma_y - 2\sqrt{\Sigma_x\Sigma_y}).
-\end{equation}
-
-\noindent In this section a method to approximate the Wasserstein distance between discrete or continuous distributions in $\mathbb{R}^2$ is presented.
-Consider a physical domain with points $x_{i}, 1 \leq i \leq n \times m$, where $n, m$ are the number of points in the $x, y$ dimensions respectively.
-Form the matrix $X$ such that $(X)_{j, k}$ is the amount transported from $x_j$ under $f$ to $x_k$ under $g$.
-Form the matrix $(C)_{j, k} = |x_j - x_k|$ as the distance between points $x_j$ and $x_k.$  With this framework, computing the WD is then equivalent to solving
-\begin{equation}
-\begin{aligned}
-& {\text{minimize}}
-& \sum_{ij}X_{ij}C_{ij} \\
-& \text{subject to}
-& X\textbf{1} = (a_1, \dotsc, a_m)^{T} \\ 
-& & \textbf{1}^{T}X = (b_1, \dotsc, b_n)\\
-& & (X)_{ij} \geq 0,
-\end{aligned}
-\label{eq: dual}
-\end{equation}
-where \textbf{1} indicates a vector of ones of appropriate length.
-Because of the linearity of the objective function, (\ref{eq: dual}) can be solved with the simplex method.
-In Figures \ref{fig: 2Ddiscreet} and \ref{fig: 2DContinuous}, the WD between discreet functions and continuous functions, respectively, in two dimensions is computed.
-
- \begin{figure}[H]
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \includegraphics[scale = .4]{2DPoints.jpg}
-  \caption{Discrete Points, WD $\approx$ 5.69}
-  \label{fig: 2Ddiscreet}
-\end{subfigure}%
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \includegraphics[scale = .4]{2DGaussians.jpg}
-  \caption{Continuous Functions, WD $\approx$ 7.12}
-  \label{fig: 2DContinuous}
-\end{subfigure}
-\caption{Computations of WD in two dimensions}
-\label{fig:1DWDCalculations}
-\end{figure}
-\noindent \textbf{Remark:} Although the continuous examples given were normal distributions, the WD can be computed between any two continuous functions.
-
-\section{Application to Biological Systems}
-\label{sec: BioSystems}
-This section applies the tools from previous sections to quantify the differences between biological models that use particles to study larger phenomenon (``micro" models), such as the growth of cancer, and those that study such phenomenon on a macroscopic scope.
-This question is of critical importance to researchers, since the micro-level models are computationally cheaper and thus providing measures of their accuracy can be very useful.
-
-\subsection{One Dimensional Case}
-For illustration, we begin by examining the growth of cancer cells. To model such a phenomenon on a macroscopic level, we study the model
-\begin{equation}
-\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u},
-\label{eq: MacroModel}
-\end{equation}
-where $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. The solution to ($\ref{eq: MacroModel}$) is the well-known Gaussian and is given by
-
-\begin{equation*}
-Macro_\rho(x) = \frac{1}{\sigma \sqrt{2\pi}}\exp\Big(\frac{-(x-\mu)^2}{2\sigma^2}\Big)
-\end{equation*}
-
-\noindent For our micro-model, we consider the model given by
-\begin{equation}
-dX_t = \sigma dB_t,
-\label{eq: MicroModel}
-\end{equation}
-where $B_t$ denotes a standard Brownian motion, with $B(0) = 0$. In order to compare our micro-model to the macro-model, we approximate (\ref{eq: MicroModel}). For $\Delta t > 0$, we have
-\begin{equation}
-\int_0^{\Delta t} \ dX_t = \sigma \int_{0}^{\Delta t} \ dB_t \implies x_{\Delta t} - x_0 = \sigma \textbf{Z}
-\end{equation}
-where $\textbf{Z} \sim \mathcal{N}(0, \Delta t)$. Using properties of variance, 
-\begin{equation}
-x_{\Delta t} = x_0 + \sigma \sqrt{\Delta t}\textbf{W}
-\label{eq: BrownianEQ}
-\end{equation} 
-where $\textbf{W} \sim \mathcal{N}(0, 1)$.
-To compute $Micro_{\rho}$, $N \in \mathbb{Z}^+$ i.i.d.
-realizations of (\ref{eq: BrownianEQ}) are computed, from which a histogram ($Micro_{\rho}$) is created over the physical domain.
-In Figures \ref{fig: rhoInitial} and \ref{fig: rhoFinal} we plot $Macro_{\rho}, Micro_{\rho}$ at the initial and final times, respectively.
-\begin{figure}[H]
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{InitialDistributions.jpg}
-  \caption{Solutions at initial time}
-  \label{fig: rhoInitial}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{FinalDistribution.jpg}
-  \caption{Solutions at final time}
-  \label{fig: rhoFinal}
-\end{subfigure}
-\caption{$N = 10^3, \ T_{final} = 1, \ \Delta t =.01.$}
-\label{fig:test}
-\end{figure}
-
-\noindent Using a standard normal PDF as the initial condition for $Macro_{\rho}$, and $\Delta x = .05$, we see in Figures \ref{fig:FittingError} and \ref{fig: FittingLogError} the decay of WD and log(WD) against number of particles, respectively.
-All WD reported are at $T_{final} = 1, \ \Delta t = .01.$
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{WDVSN1D.jpg}
-  \caption{Fit WD and N}
-  \label{fig:FittingError}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{LogWDVSN1D.jpg}
-  \caption{Fit Log(WD) and Log(N)}
-  \label{fig: FittingLogError}
-\end{subfigure}
-\caption{Decay of WD with N}
-\label{fig:test}
-\end{figure}
-
-\noindent We now explore how changing $\Delta x$, for a fixed $N$, changes the WD.
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[scale = .4]{WDVSDX.jpg}
-\end{center}
-\caption{WD vs $\Delta x$.}
-\end{figure}
-
-\noindent We conclude from Figure \ref{fig:FittingError}  that the error between Macro$_\rho$ and Micro$_\rho$ follows the relationship 
-\begin{equation*}
-\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
-\end{equation*}
- where $C \approx 3.33, \ \alpha \approx 0.44 $.
-When computing the linear fit  $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$.
-We thus conjecture that the Wasserstein distance between Micro$_\rho$ and Macro$_\rho$ decays at a rate of $1/\sqrt{N}$.
-\begin{theorem}
-$\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
-\end{theorem}
-\begin{proof}
-Denote Macro$_\rho$ by $\rho_{mi}$, Micro$_\rho$ by $\rho_{ma}$. Then, for some $\phi \in C_b(\mathbb{R})$
-\begin{equation*}
-\langle \rho_{mi}, \phi \rangle = \int_{\mathbb{R}} \rho_{mi} \phi(x) \ dx = \frac{1}{N}\sum_x \phi(x_i(t)).
-\end{equation*}
-By the Law of Large Numbers, 
-\begin{equation*}
-\frac{1}{N}\sum_x \phi(x_i(t)) \xrightarrow{N \rightarrow \infty} \int_{\mathbb{R}} \phi(x)\rho_{mi}(x) \ dx + \frac{1}{\sqrt{N}}\text{std}(\phi(x))
-\end{equation*}
-where $\text{std}(\phi(x))$ denotes the standard deviation of $\phi$.
-\end{proof}
-
-\noindent We now consider a model with a source term, meaning the macro-level model is given by
-
-\begin{equation}
-\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u} + \rho(1-\rho),
-\label{eq: MacroModel2}
-\end{equation}
-
-To calculate $\rho$Micro, $N$ realizations of (\ref{eq: BrownianEQ}) are computed.
-At each time-step, a new particle is created with probability $\Delta t(1-\rho)$ and destroyed with probability .
-A histogram ($Micro_{\rho}$) is then
-computed over the physical domain.
-
-\subsection{Two Dimensional Case}
-
-In two dimensions, the equation for the macro model becomes 
+\section{Shane's Part}
+In two dimensions, one can model the macroscopic behavior of the system by
 \begin{equation}
@@
 \end{equation}
-where $\rho(t,x,y)$ denotes the density of cancer cell at time $t$ at location $(x,y).$ Similarly, (\ref{eq: BrownianEQ}) becomes 
+where $\rho(t,x,y)$ is the density of cancer cell at time $t$ at location $(x,y),$ and $D$ is the diffusion coefficient. One can model the microscopic behavior of the system by
+
 \begin{equation}
@@
 \end{equation}
-where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.
+where $\sigma \in \mathbb{R}$, and $B_t$ denotes a standard two-dimensional Brownian motion, with $B(0) = (0,0)$. Thus,
 
-\section{Conclusion}
-\label{sec: Conclusion}
+\begin{equation}
+\int_0^{\Delta t} \ dX_t = \sigma \int_{0}^{\Delta t} \ dB_t \implies x_{\Delta t} - x_0 = \sigma \textbf{Z}
+\end{equation}
+where $\textbf{Z} \sim \mathcal{N}_2(0, \Delta t \mathbb{I}_2)$, where $\mathbb{I}_p$is the $p \times p$ identity matrix. We conclude that
+\begin{equation}
+x_{\Delta t} = x_0 + \sigma \sqrt{\Delta t}\textbf{W},
+\label{eq: BrownianEQ}
+\end{equation} 
+where $\textbf{W} \sim \mathcal{N}_2(0, \mathbb{I}_2).$ In Figures \ref{fig:macroPlot} and \ref{fig:microPlot} we plot the level curves of the solution to (\ref{eq: Macro2D}) and (\ref{eq: BrownianEQ}) respectively.
+In Figure \ref{fig: WD}, we plot the WD between Micro and Macro as a function of $t$, where we use $N = 10^3, \Delta t = .01, t = 1, \Delta x = \Delta y = .5.$
 
-\section{Definition: More for My Use}
-\begin{itemize}
-\item Let $X$ be a set. An algebra on $X$ is a collection $\Sigma \subseteq 2^X$ that is closed under finite set operations (complement, union, and intersection). A 
- $\sigma-$algebra on $X$ is a collection $\Sigma \subseteq 2^X$ that is closed under countably many such operations. The intersection of a collection of $\sigma-$algebras is a $\sigma-$algebra. 
- \item A \textbf{Borel set} in a topological space $(X, \mathscr{T})$ is a set that can be made from open sets $U \in \mathscr{T}$ by countable union, countable intersection, or relative complement.
-The collection of all Borel sets on $X$ forms a $\sigma-$algebra, known as the Borel $\sigma-$algebra, and is denoted by $\mathcal{B}(X)$.
- \item Let $X, Y$ be two topological spaces.
-A function $f: X \rightarrow Y$ is called a Borel map if 1) $f^{-1}(A)$, A open, is a Borel subset, 2) $f^{-1}(B)$, B closed, is a Borel subset, and 3) $f^{-1}(C)$, C Borel, is a Borel subset.
- \item Given a measurable space (X, $\Sigma$) and a measure $\mu$ on that space, a set $A \subset X$ in $\Sigma$ is called an \textbf{atom} if $\mu(A) > 0$ and for any measurable subset $B \subset A$ with $\mu(B) < \mu(A)$, the set B has measure zero.
-A measure that has no atoms is called \textbf{non-atomic}.
-\end{itemize}
-
-Informal sources:
-\begin{itemize}
-\item (1) http://www.math.umd.edu/~yanir/OT/AmbrosioGigliDec2011.pdf
-\item (2) http://www.math.tohoku.ac.jp/~aida/workshop/H23-Daisympo/kuwada.pdf
-\end{itemize}
-\begin{thebibliography}{9}% maximum number of references (for label width)
-\bibitem{CVillani}
-Villani, C., Optimal transport, old and new, June 13, 2008, http://cedricvillani.org/wp-content/uploads/2012/08/preprint-1.pdf
-
-\bibitem {AN}
-Ambrosio, L., Gigli, N. A user's guide to optimal transport.  \url{http://www.math.umd.edu/~yanir/OT/AmbrosioGigliDec2011.pdf}.
-
-\bibitem{Downson}
-Dowson, D.C., B.V Landau, The Frechet Distance between Multivariate Normal Distributions, J. Multivariate Analysis 12, 450-455 (1982).
-
-\bibitem{P}  
-Prokhorov, Yu. V., Convergence of random processes and limit theorems in probability theory. (Russian) Teor. Veroyatnost. i Primenen. 1 (1956), 177-238.
-
-\end{thebibliography}
+\begin{figure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[scale = .45]{MacroPlot.jpg}
+\caption{Macro Solution}
+\label{fig:macroPlot}
+\end{subfigure}%
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[scale = .45]{MicroPlot.jpg}
+\caption{Micro Solution}
+\label{fig:microPlot}
+\end{subfigure}\\[1ex]
+\begin{subfigure}{\linewidth}
+\centering
+\includegraphics[scale = .45]{WD.jpg}
+  \caption{Wasserstein Distance}
+\label{fig: WD}
+\end{subfigure}
+\caption{2D Figures}
+\label{fig:test}
+\end{figure}
 
 \end{document}
-
-% - Release $Name:  $ -

### 2016-08-18 model_2D/lib/Micro.m

@@ -5,23 +5,6 @@
 %  + 'birth process' with intensity 1-?.
 % 
 
-%  function [rho] = KPP_Micro(rhoIC,x,D,dt,T,N,seedNum)
-% 
-% Solve the KPP equation as a particle system:
-%    dX_t = ?dB_t
-%  + 'birth process' with intensity 1-?.
-
-%     x= linspace(-2,2,10);
-%     y = x;
-%     D= 100;
-%     T = 1000;
-%     dt = .1;
-%     N = 10^3;
-%     mu1 = 1;
-%     mu2 = 0;
-%     var1 = 1;
-%     var2 = 2;
-
     [xmesh,ymesh] = meshgrid(x,y);
     rho = zeros(numel(x),numel(y),floor(T/dt + .5));
     
@@ -60,14 +43,6 @@
 %         toc
     
     end
-    
-%     figure;
-%     for j = 1:floor(T/dt + .5)
-%         surf(rho(:,:,j)); grid on;
-%         xlabel('x'); ylabel('y'); zlabel('f(x,y)');
-%         pause(.1);
-%     end
-%    
 
  end
 

### 2016-08-18 model_2D/lib/TwoDStochastic.m

@@ -6,19 +6,19 @@
 clc; clear all; close all;
 
 %% Parameters
-D = 10;                                % diffusion coefficients
+D = 1;                                % diffusion coefficients
 %% Parameters micro
-N = 10^4;                               % number of cells (initially)
+N = 10^3;                               % number of cells (initially)
 Mass_rhoIC = 2;
 seedNum = 1;
 % domain
-dx = .8;
-dy = .8;
-x = -2:dx:2;
-y = -2:dy:2;
+dx = .25;
+dy = .25;
+x = -5:dx:5;
+y = -5:dy:5;
 % time
-dt = 10^-2;
-T =   1;
+dt = .5;
+T =   5;
 t = 0:dt:T;
 
 % Initial Conditiion
@@ -29,48 +29,83 @@ var2 = 1;
 shouldPlot = 0;
 
 
+
+for q = 1:length(t)
 % Call Micro, Macro
 %------------------
-% rhoMicro = KPP_Micro_2D(x,y,D,dt,T,N,seedNum,Mass_rhoIC,mu1,mu2,var1,var2);
 rhoMicro = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2);
 [rhoDir,rho] = Macro(x,y,D,dt,T,mu1,mu2,var1,var2,true);
 
+rhoMicro(:,:,q) = rhoMicro(:,:,q)/sum(sum(rhoMicro(:,:,q)));
+rho(:,:,q) = rho(:,:,q)/sum(sum(rho(:,:,q)));
 
 % Compute speed of traveling waves
 %---------------------------------
-addpath('lib')
-c = .2; % threshold for computing contours.
-for j = 1:floor(T/dt + .5)
-    distanceMacro(j) = ComputeVelocity(x,y,rho(:,:,j),c);
-end
-Fit = polyfit(t(2:end),distanceMacro,1);
-figure; 
-plot(t(2:end),distanceMacro,t(2:end),polyval(Fit,t(2:end)))
-xlabel('Time'); ylabel('Velocity');
+% addpath('lib')
+% c = .2; % threshold for computing contours.
+% for j = 1:floor(T/dt + .5)
+%     distanceMacro(j) = ComputeVelocity(x,y,rho(:,:,j),c);
+% end
+% Fit = polyfit(t(2:end),distanceMacro,1);
+% figure; 
+% plot(t(2:end),distanceMacro,t(2:end),polyval(Fit,t(2:end)))
[…]

### 2016-08-18 Paper/Manuscript/Shane's Part/ShaneManuscript.tex
New Data

@@
 where $\textbf{W} \sim \mathcal{N}_2(0, \mathbb{I}_2).$ In Figures \ref{fig:macroPlot} and \ref{fig:microPlot} we plot the level curves of the solution to (\ref{eq: Macro2D}) and (\ref{eq: BrownianEQ}) respectively.
-In Figure \ref{fig: WD}, we plot the WD between Micro and Macro as a function of $t$, where we use $N = 10^3, \Delta t = .01, t = 1, \Delta x = \Delta y = .5.$
+We use $N = 10^3, L = M = 5, \Delta x = \Delta y = .2$ at $t = 0,$ where $x \in [-L, L], \ y \in [-M, M]$.
+In Figure \ref{fig: WD}, we plot the WD between Micro and Macro as a function of $t$, where we use $N = 10^3,\ \Delta t = .01, \ t = 1, \\ L = M  = 2, \Delta x = \Delta y = .5$
 
 \begin{figure}
-\begin{subfigure}{.45\linewidth}
+\begin{subfigure}{.5\linewidth}
 \centering
@@
 \end{subfigure}%
-\begin{subfigure}{.45\linewidth}
+\begin{subfigure}{.5\linewidth}
 \centering

### 2016-08-18 model_2D/lib/TwoDStochastic.m

@@ -6,19 +6,19 @@
 clc; clear all; close all;
 
 %% Parameters
-D = 1;                                % diffusion coefficients
+D = .1;                                % diffusion coefficients
 %% Parameters micro
 N = 10^3;                               % number of cells (initially)
 Mass_rhoIC = 2;
 seedNum = 1;
 % domain
-dx = .25;
-dy = .25;
-x = -5:dx:5;
-y = -5:dy:5;
+dx = .5;
+dy = .5;
+x = -2:dx:2;
+y = -2:dy:2;
 % time
-dt = .5;
-T =   5;
+dt = .1;
+T =   20;
 t = 0:dt:T;
 
 % Initial Conditiion
@@ -29,12 +29,13 @@ var2 = 1;
 shouldPlot = 0;
 
 
+rhoMicro = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2);
+[rhoDir,rho] = Macro(x,y,D,dt,T,mu1,mu2,var1,var2,true);
+ 
 
 for q = 1:length(t)
 % Call Micro, Macro
 %------------------
-rhoMicro = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2);
-[rhoDir,rho] = Macro(x,y,D,dt,T,mu1,mu2,var1,var2,true);
 
 rhoMicro(:,:,q) = rhoMicro(:,:,q)/sum(sum(rhoMicro(:,:,q)));
 rho(:,:,q) = rho(:,:,q)/sum(sum(rho(:,:,q)));

### 2016-08-23 Final_Manuscript/Shane's Part/ShaneManuscript.tex
8/23

--- 2016-08-18 Paper/Manuscript/Shane's Part/ShaneManuscript.tex
@@
 % do not change this formatting 
-\documentclass[10pt]{article}
+\documentclass[10pt, margins = .5in]{article}
 \parindent=0pt \parskip=8pt
@@
 \newcommand*{\QEDA}{\hfill\ensuremath{\blacksquare}}%
-\usepackage[margin=1in]{geometry}
+\usepackage[margin=.5in]{geometry}
 \usepackage{xcolor}
@@
 \end{equation}
-where $\sigma \in \mathbb{R}$, and $B_t$ denotes a standard two-dimensional Brownian motion, with $B(0) = (0,0)$. Thus,
+where $\sigma = \sqrt{2D}$, and $B_t$ denotes a standard two-dimensional Brownian motion, with $B(0) = (0,0)$. Thus,
 
@@
 \end{equation}
-where $\textbf{Z} \sim \mathcal{N}_2(0, \Delta t \mathbb{I}_2)$, where $\mathbb{I}_p$is the $p \times p$ identity matrix. We conclude that
+where $\textbf{Z} \sim \mathcal{N}_2((0, 0), \Delta t \mathbb{I}_2)$, where $\mathbb{I}_p$is the $p \times p$ identity matrix. We conclude that
 \begin{equation}
@@
 \end{equation} 
-where $\textbf{W} \sim \mathcal{N}_2(0, \mathbb{I}_2).$ In Figures \ref{fig:macroPlot} and \ref{fig:microPlot} we plot the level curves of the solution to (\ref{eq: Macro2D}) and (\ref{eq: BrownianEQ}) respectively.
-In Figure \ref{fig: WD}, we plot the WD between Micro and Macro as a function of $t$, where we use $N = 10^3, \Delta t = .01, t = 1, \Delta x = \Delta y = .5.$
+where $\textbf{W} \sim \mathcal{N}_2((0,0), \mathbb{I}_2).$ In Figures \ref{fig:macroPlot} and \ref{fig:microPlot} we plot the level curves of the solution to (\ref{eq: Macro2D}) and (\ref{eq: BrownianEQ}) respectively.
+We use $N = 10^4, L = M = 5, \Delta x = \Delta y = .2$ at $t = 0,$ where $x \in [-L, L], \ y \in [-M, M]$.
+In Figure \ref{fig: WD}, we plot the Wasserstein Distance between the solutions of (\ref{eq: Macro2D}) and (\ref{eq: MicroModel2}) as a function of $t$, where we use $N = 10^3,\ \Delta t = .05, \ t = 20, L = M  = 2, \ \Delta x = \Delta y = .5.$
 
-\begin{figure}
-\begin{subfigure}{.45\linewidth}
+\begin{figure}[H]
+\begin{subfigure}{.5\linewidth}
 \centering
-\includegraphics[scale = .45]{MacroPlot.jpg}
+\includegraphics[scale = .40]{MacroPlot.jpg}
 \caption{Macro Solution}
@@
 \end{subfigure}%
-\begin{subfigure}{.45\linewidth}
+\begin{subfigure}{.40\linewidth}
 \centering
-\includegraphics[scale = .45]{MicroPlot.jpg}
+\includegraphics[scale = .40]{MicroPlot.jpg}
 \caption{Micro Solution}
@@
 \end{subfigure}\\[1ex]
-\begin{subfigure}{\linewidth}
 \centering
+\begin{subfigure}{.45\linewidth}
 \includegraphics[scale = .45]{WD.jpg}
@@
 \end{subfigure}
-\caption{2D Figures}
+\caption{2D WD Data}
 \label{fig:test}

### 2016-08-23 OriginsEssay.tex
8/23

\documentclass[8 pt]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{fancyhdr}
\usepackage{wrapfig}
\usepackage[margin=.7in]{geometry}
\usepackage{amssymb,amsmath,amsthm,mathrsfs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{mdframed}

\pagestyle{fancy}
\fancyhf{}
    \rhead{Origins Project Research Proposal}
\lhead{Shane M. Lubold}
\linespread{1}

\begin{document}

As our computing power inceases, questions that once seemed impossible to analyze become accessible. Faster computation has created new discoveries in economics, computer science, and engineering, but nowhere are these discoveries more important than in medicine. Biological processes, such as the growth of cancer, the spread of disease, or the evolution of species, are at the forefront of modern science. Answering these questions is crucial to our progress as a species. Reducing the time needed to model the growth of cancer, for instance, would translate into quicker diagnoses and faster, more effective treatments. Working under Dr. S\'{e}bastien Motsch in the School of Mathematical and Statistical Sciences, I plan to focus on three aspects of computational biology.

 Dr.\ Motsch and I are currently deriving an alternative to the traditional approach to modeling the behavior of biological systems. The traditional approach is to use differential equations to model such systems. For instance, 
a biologist interested in modeling cancer growth would try to solve:

\begin{equation}
\frac{\partial f}{\partial t} = D\Big(\frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}\Big)+ f(1-f).
\label{eq: Macro}
\end{equation}

To solve such an equation, one discretizes the physical space (corresponding to the $x$ and $y$ variables) and the temporal space (corresponding to the $t$ variable). This approach requires the ratio $\frac{\Delta t}{\Delta x^2}$ to be very small in order for the solution to be accurate. This results in a computationally expensive problem. Furthermore, as the physical space gets larger, the computational time required increases exponentially. For problems of even moderate size, this approach is therefore impracticle.\\

\begin{wrapfigure}{L}{0.4\textwidth}
  \begin{center}
   \includegraphics[width=0.4\textwidth]{OriginsPic.jpg}
  \end{center}
  \caption{Comparison of Approaches}
\end{wrapfigure}

\noindent The second approach is to model the behavior of the physical system using particles. Instead of solving a complicated differential equation, one begins with a collection of particles, with $X_i^{m}$ denoting the $i$-th particle at time $m.$ One then computes the position of these particles at a future time with  
\begin{equation}
x_i^{m+1} = x_i^{m} + C\textbf{W},
\label{eq: Micro}
\end{equation}
where $C$ is a constant and $\textbf{W} \sim \mathcal{N}(0, 1).$ One simply needs to draw normally distributed numbers in order to approximate the behavior of the system. One can prove mathematically that these particles converge in distribution to the equation of the differential equation. In Figure 1,
the solutions from (\ref{eq: Macro}) and (\ref{eq: Micro}) are plotted.
This approach is computationally faster and places no restriction on the physical and temportal discretization, yet there is a problem. There has been little research done that computes the accuracy of this second scheme in relation to the first. Can one trust the answers given by the second approach? \\
The goal of my and Dr. Motsch's research is to prove that the second approach provides an accurate approximation to the system one is considering. We do this using both numerical and pure mathematics. What we show is that researchers in fields as diverse as biology, chemistry, medecine, can faithfully rely on the approximations coming from the second approach. The end result is that researchers in these fields can now solve probems that once seemed inaccesible a few years ago because essentially their computing power has been increased significnatly.

The answer is that before now, there was very little way to compare the accuracy of the second approach to the first approach. Some ways existed, but they were complicated and required advanced mathematics to fully understand. What Dr. Motsch and I are doing is simplifying the mathematics and deriving simple algorithms that researchers in fields as diverse as biology, chemistry, medecine, etc. can use in their research. What we are doing is allowing these researchers to avoid the complicated mathematics and use the second approach, knowing that this approach is ``close enough'' to the first approach to not be sacrificing accuracy.

Our first paper is set to be submitted in October. This paper demonstrates that it is possible to use this second approach in approximating biological processes that can be approximated through differential equations. This means that most problems in biology can now be solved through the second approach. Our research indicates that the second approach is 4 times faster than the first approach.

This project lies at the intersection of evolutionary and computational biology, pure mathematics, applied mathematics, and statistics. Dr. Motsch and I are qualified to carry out this project for several reasons. Dr. Motsch received his PhD in applied mathematics with an emphasis in animal cognition and mathematical biology, while I am undergraduate student in pure mathematics intending to obtain a PhD in statistics. Having the opportunity to conduct research as a Origins Scholarship Recipient will allow me to focus solely on my research and finish a second paper with Dr. Motsch. Having three papers on my CV when I apply to PhD programs will be incredibly beneficial.

The timeline for our project is as follows. In August and September, Dr. Motsch and I are completing work on our first paper, which examines some appliations of the Wasserstein distance to studying the growth of cancer cells. This paper will be submitted at the beginning of October. We will then begin working on a second paper,

\end{document}

### 2016-08-23 model_1D/Micro_Macro.m

@@ -13,9 +13,9 @@ D = 2/3;
 dt = .1;
 T  = 2;
 dx = .1;
-L  = 10;
+L  = 3;
 method_Kpp = 2;                           % 1: explicit, 2: Cranck-Nicholson
-Ninit  = 1e6;                           % number of particles initially
+Ninit  = 1e2;                           % number of particles initially
 seedNum = 3;
 % Initial condition
 x = -L:dx:L;

### 2016-08-23 model_2D/lib/TwoDStochastic.m

@@ -6,19 +6,19 @@
 clc; clear all; close all;
 
 %% Parameters
-D = .1;                                % diffusion coefficients
+D = 5;                                % diffusion coefficients
 %% Parameters micro
-N = 10^3;                               % number of cells (initially)
+N = 10^4;                               % number of cells (initially)
 Mass_rhoIC = 2;
 seedNum = 1;
 % domain
 dx = .5;
 dy = .5;
-x = -2:dx:2;
-y = -2:dy:2;
+x = -5:dx:5;
+y = -5:dy:5;
 % time
 dt = .1;
-T =   20;
+T =   5;
 t = 0:dt:T;
 
 % Initial Conditiion

### 2016-10-18 FinalOrigins.tex
10/18

\documentclass[8 pt, leqno]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{fancyhdr}
\usepackage{wrapfig}
\usepackage[margin=.75in]{geometry}
\usepackage{amssymb,amsmath,amsthm,mathrsfs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{pgfplots}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\pgfplotsset{
  compat=1.9,
  unit code/.code 2 args={\si{#1#2}} % from manual, for using siunitx to typeset units
}
\usepackage{mdframed}
\usepackage{inputenc}
\usepackage{siunitx}
\usepackage{siunitx}
\usepgfplotslibrary{groupplots,units}
\newlength\figureheight 
\newlength\figurewidth 
\setlength\figureheight{0.4\textwidth}
\setlength\figurewidth{0.35\textwidth}
\usepackage{tikz}
\setcounter{section}{2}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{multicol}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[CO]{A Novel Approach To Modeling Complex Systems with Brownian Motion}
\linespread{1.5}

%\title{}
%\date{}
%\author{}
\begin{document}
%\maketitle

Modeling complex systems, from the growth of cancer cells and bacteria to the evolution of a species, is critical to our understanding of the physical world and our place in it. To comprehend our origins, it is essential to accurately model the dynamics driving the creation of life, and to comprehend our future, it is essential to have accurate predictive models. Such modeling is complicated when the dynamics behind these complex systems is random, or stochastic.\\
\indent Current models describing biological processes are deterministic, meaning they do not accurately reflect the randomness of complex systems over long periods of time. Dr. Motsch, a professor of applied biology and statistics in the School of Mathematical and Statistical Sciences, and I will study the same complex systems with the use of Brownian motion in order to obtain more accurate predictions of the behavior of stochastic complex systems. Brownian motion, originally studied by the botanist Robert Brown in the 19th century and popularized by Einstein in the early 20th century, describes motion that behaves stochastically. For instance, if a particle suspended in water were acted upon randomly by other particles, its behavior would be described as Brownian. My research with Dr. Motsch will seek to demonstrate that Brownian motion will accurately model stochastically complex systems and allow us to invent frameworks that derive meaningful insights into the future.\\
\indent In order to quantify the accuracy of our approach, we will be using a tool in mathematics called the Wasserstein Distance, which quantifies the distance between functions or data sets, and thereby provides us insight into the improvement that our approach to modeling stochastic systems offers. We aim to collaborate with
Dr. Pedro Lowenstein, a professor of Neurosurgery at the University of Michigan to obtain data on cancer growth in a mouse, and with this data, we will use the Wasserstein Distance to compute the distance between the true biological data, the predictions made by the Brownian motion approach, and the predictions made by the deterministic approach. We seek to understand how the number of particles used in the Brownian motion approach affects the accuracy of our approximation of the biological data. More specifically, we will propose an algorithm that will determine the number of particles necessary to get within a desired error using the Brownian motion approach. \\
\indent During a summer research project, completed under Dr. Anne Gelb in 2014, I studied the use of statistical algorithms to improve Fourier edge detection in the presence of corrupted data. Specifically, Dr. Gelb and I derived an algorithm for reconstructing MRI images that is 166$\%$ more accurate than previous methods available to the medical community. By working with Dr. Gelb, I developed an interest in using statistical tools to study relevant problems in biology. During the summer of 2016 I participated in two summer research programs that provided me with the interest and background necessary to successfully complete my Origins Project research. My first project was at the University of California, Santa Barbara where I took a series of courses on differential equations in random media, biological modeling of living systems, and statistical approximations of biological processes. Through conversations with professors and graduate students from Princeton, Stanford, UC Berkeley and others, I learned about cutting-edge research at the intersection of biology and statistics and will apply these techniques in my Origins Project research.\\
\indent In my second project, I was selected as one of three from over 300 undergraduates to participate in a research experience at San Diego State University. There I studied the convergence behavior of statistical algorithms used in linear regression models. Specifically, my advisor and I proposed the first algorithm for concrete convergence bounds for the error of a Gibbs sampler in a Bayesian linear regression model. These results will be submitted for publication in October. This research experience taught me the tools to complete statistical convergence analysis, and I will use similar tools to study the convergence behavior of the approximation to complex systems using Brownian motion. \\
\indent To successfully complete this research, I will need an advanced grasp of real analysis, probability theory, and numerical analysis. . I have taken a course on numerical methods with Dr. Motsch, and I have enrolled in graduate courses in real analysis, distribution theory, and stochastic processes. In these courses I will obtain the tools necessary to analyze papers relating to my work with Dr Motsch, as well as make meaningful contributions to the field. In the Spring I will finish the graduate sequence in real analysis and probability theory, creating a solid theoretical foundation that I will use in my research.\\
\indent Developing accurate statistical algorithms for understanding stochastic complex systems is one of the largest problems in modern biological sciences, and the opportunity to study this type of problem has been my main focus during my undergraduate career. In addition to contributing to the field of computational biology, I hope to motivate theoretical mathematicians and statisticians to solve problems of interest in applied fields, leading to more interactions and discussions between the previously distant groups of researchers. Indeed, the Wasserstein distance might also be successfully applied to problems in computer science and chemistry. In addition to significantly adding to my knowledge in statistics and computational mathematics, I hope this research project will start a larger discourse on the use of statistical algorithms, specifically Brownian motion, in biology, chemistry, and other applied fields. \\
\indent The Origins Project represents an avenue towards galvanizing an interest among pure mathematicians and statisticians in using theoretical tools to study problems of interest in biology. With the Origins Project’s Undergraduate Research Scholarship, I will have the opportunity to demonstrate the utility of mathematics and statistics in computational biology. By improving the accuracy of approximations in biology, which studies the development of infectious diseases and our place in the universe, my research will shed light on the origin and future of humanity.

\end{document}

### 2016-10-18 FinalOrigins_v2.tex
10/18

--- 2016-10-18 FinalOrigins.tex
@@
 \usepackage{inputenc}
-\usepackage{siunitx}
-\usepackage{siunitx}
+%\usepackage{siunitx}
+%\usepackage{siunitx}
 \usepgfplotslibrary{groupplots,units}
@@
 \fancyhf{}
-\fancyhead[CO]{A Novel Approach To Modeling Complex Systems with Brownian Motion}
+\fancyhead[CO]{A Novel Approach To Modeling Random Complex Systems with Brownian-type Motion}
 \linespread{1.5}
@@
 To comprehend our origins, it is essential to accurately model the dynamics driving the creation of life, and to comprehend our future, it is essential to have accurate predictive models.
-Such modeling is complicated when the dynamics behind these complex systems is random, or stochastic.\\
-\indent Current models describing biological processes are deterministic, meaning they do not accurately reflect the randomness of complex systems over long periods of time.
-Dr.
-Motsch, a professor of applied biology and statistics in the School of Mathematical and Statistical Sciences, and I will study the same complex systems with the use of Brownian motion in order to obtain more accurate predictions of the behavior of stochastic complex systems.
+Such modeling is complicated when the dynamics behind these complex systems is random, also referred to as {\it stochastic}.\\
+\indent Current models of biological processes focus on  deterministic behavior, but randomness is essential to determine how complex systems behave over long periods of time.
+Dr.\ Motsch, a professor of applied biology and statistics in the School of Mathematical and Statistical Sciences, and I will study the same complex systems with the use of Brownian motion in order to obtain more accurate predictions of their behavior.
 Brownian motion, originally studied by the botanist Robert Brown in the 19th century and popularized by Einstein in the early 20th century, describes motion that behaves stochastically.
 For instance, if a particle suspended in water were acted upon randomly by other particles, its behavior would be described as Brownian.
-My research with Dr.
-Motsch will seek to demonstrate that Brownian motion will accurately model stochastically complex systems and allow us to invent frameworks that derive meaningful insights into the future.\\
-\indent In order to quantify the accuracy of our approach, we will be using a tool in mathematics called the Wasserstein Distance, which quantifies the distance between functions or data sets, and thereby provides us insight into the improvement that our approach to modeling stochastic systems offers.
-We aim to collaborate with
-Dr.
-Pedro Lowenstein, a professor of Neurosurgery at the University of Michigan to obtain data on cancer growth in a mouse, and with this data, we will use the Wasserstein Distance to compute the distance between the true biological data, the predictions made by the Brownian motion approach, and the predictions made by the deterministic approach.
+In the figure below we plot Brownian particles modeling cancer cell growth on top of blood vessels.
+My research with Dr.\ Motsch will seek to demonstrate that Brownian-type motion will accurately model complex systems and allow us to invent mathematical frameworks that derive meaningful insights into our past and the future.\\
+\begin{wrapfigure}[11]{R}{0.4\textwidth}
+    \includegraphics[width=0.4\textwidth]{Graph.pdf}
+  \caption*{}
+\end{wrapfigure} 
+\indent In order to quantify the accuracy of our approach, we will be using a tool in mathematics called the Wasserstein Distance, which measures the distance between functions or data sets, and thereby provides us insight into the improvement that our approach to modeling stochastic systems offers.
+We will collaborate with
+Dr.\ Pedro Lowenstein, a professor of Neurosurgery at the University of Michigan, to obtain data on cancer growth in mice.
+We will use the Wasserstein Distance to compare the experimental data, the predictions made by the Brownian motion approach, and the predictions made by the deterministic approach.
 We seek to understand how the number of particles used in the Brownian motion approach affects the accuracy of our approximation of the biological data.
 More specifically, we will propose an algorithm that will determine the number of particles necessary to get within a desired error using the Brownian motion approach.
-\\
-\indent During a summer research project, completed under Dr.
-Anne Gelb in 2014, I studied the use of statistical algorithms to improve Fourier edge detection in the presence of corrupted data.
-Specifically, Dr.
-Gelb and I derived an algorithm for reconstructing MRI images that is 166$\%$ more accurate than previous methods available to the medical community.
-By working with Dr.
-Gelb, I developed an interest in using statistical tools to study relevant problems in biology.
+
+\indent During a summer research project, completed under Dr.\ Anne Gelb, I studied the use of statistical algorithms to improve Fourier edge detection in the presence of corrupted data.
+Specifically, Dr.\ Gelb and I derived an algorithm for reconstructing MRI images that is twice as accurate as previous methods available to the medical community.
+By working with Dr.\ Gelb, I developed an interest in using statistical tools to study relevant problems in biology.
 During the summer of 2016 I participated in two summer research programs that provided me with the interest and background necessary to successfully complete my Origins Project research.
-My first project was at the University of California, Santa Barbara where I took a series of courses on differential equations in random media, biological modeling of living systems, and statistical approximations of biological processes.
+My first project was at the University of California, Santa Barbara, where I took a series of graduate courses on differential equations in random media, biological modeling of living systems, and statistical approximations of biological processes.
 Through conversations with professors and graduate students from Princeton, Stanford, UC Berkeley and others, I learned about cutting-edge research at the intersection of biology and statistics and will apply these techniques in my Origins Project research.\\
@@
 There I studied the convergence behavior of statistical algorithms used in linear regression models.
-Specifically, my advisor and I proposed the first algorithm for concrete convergence bounds for the error of a Gibbs sampler in a Bayesian linear regression model.
+Specifically, my advisor Dr.\ Rom\'{a}n and I proposed the first algorithm for concrete convergence bounds for the error of a Gibbs sampler in a Bayesian linear regression model.
 These results will be submitted for publication in October.
-This research experience taught me the tools to complete statistical convergence analysis, and I will use similar tools to study the convergence behavior of the approximation to complex systems using Brownian motion.
+This research experience taught me the tools to complete statistical convergence analysis, and I will use similar tools to study the convergence behavior of the approximation to complex systems using Brownian-type motion.
 \\
 \indent To successfully complete this research, I will need an advanced grasp of real analysis, probability theory, and numerical analysis.
-.
 I have taken a course on numerical methods with Dr.
@@
 \indent Developing accurate statistical algorithms for understanding stochastic complex systems is one of the largest problems in modern biological sciences, and the opportunity to study this type of problem has been my main focus during my undergraduate career.
-In addition to contributing to the field of computational biology, I hope to motivate theoretical mathematicians and statisticians to solve problems of interest in applied fields, leading to more interactions and discussions between the previously distant groups of researchers.
+In addition to contributing to the field of computational biology, I hope to motivate theoretical mathematicians and statisticians to solve problems of interest in applied fields, bridging the gap between the two communities.
 Indeed, the Wasserstein distance might also be successfully applied to problems in computer science and chemistry.
@@
 With the Origins Project’s Undergraduate Research Scholarship, I will have the opportunity to demonstrate the utility of mathematics and statistics in computational biology.
-By improving the accuracy of approximations in biology, which studies the development of infectious diseases and our place in the universe, my research will shed light on the origin and future of humanity.
+By capturing the stochastic behavior of biological processes that govern the development of infectious diseases and our place in the universe, my research will shed light on the origin and future of humanity.
+
+% { (alternative: {\color[rgb]{.8,0,0} By scrutinizing the role of stochastic behavior in biological processes that govern... })
 

### 2016-10-18 OriginsEssay.tex
10/18

@@
-\documentclass[8 pt]{article}
+\documentclass[8 pt, leqno]{article}
 \usepackage[english]{babel}
@@
 \usepackage{wrapfig}
-\usepackage[margin=.7in]{geometry}
+\usepackage[margin=.75in]{geometry}
 \usepackage{amssymb,amsmath,amsthm,mathrsfs}
@@
 \usepackage{caption}
+\usepackage{pgfplots}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\pgfplotsset{
+  compat=1.9,
+  unit code/.code 2 args={\si{#1#2}} % from manual, for using siunitx to typeset units
+}
 \usepackage{mdframed}
+\usepackage{inputenc}
+\usepackage{siunitx}
+\usepackage{siunitx}
+\usepgfplotslibrary{groupplots,units}
+\newlength\figureheight 
+\newlength\figurewidth 
+\setlength\figureheight{0.4\textwidth}
+\setlength\figurewidth{0.35\textwidth}
+\usepackage{tikz}
+\setcounter{section}{2}
+\usepackage[hidelinks]{hyperref}
+\usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{multicol}
 
 \pagestyle{fancy}
-\fancyhf{}
-    \rhead{Origins Project Research Proposal}
-\lhead{Shane M. Lubold}
-\linespread{1}
+\fancyhf{A Novel Approach to Modeling Complex Systems Through Brownian Motion}
+    \rhead{}
+\lhead{}
+\linespread{1.5}
 
+%\title{}
+%\date{}
+%\author{}
 \begin{document}
+%\maketitle
 
-As our computing power inceases, questions that once seemed impossible to analyze become accessible.
-Faster computation has created new discoveries in economics, computer science, and engineering, but nowhere are these discoveries more important than in medicine.
-Biological processes, such as the growth of cancer, the spread of disease, or the evolution of species, are at the forefront of modern science.
-Answering these questions is crucial to our progress as a species.
-Reducing the time needed to model the growth of cancer, for instance, would translate into quicker diagnoses and faster, more effective treatments.
-Working under Dr.
-S\'{e}bastien Motsch in the School of Mathematical and Statistical Sciences, I plan to focus on three aspects of computational biology.
+Modeling complex systems, from the growth of cancer cells and bacteria to the evolution of a species, is critical to our understanding of the physical world and our place in it.
+To comprehend our origins, it is essential to accurately model the dynamics driving the creation of life, and to comprehend our future, it is essential to have accurate predictive models.
+Such modeling is complicated when the dynamics behind these complex systems is random, or stochastic.\\
+\indent Current models describing biological processes are deterministic, meaning they do not accurately reflect the randomness of complex systems over long periods of time.
+Dr.
+Motsch, a professor of applied biology and statistics in the School of Mathematical and Statistical Sciences, and I will study the same complex systems with the use of Brownian motion in order to obtain more accurate predictions of the behavior of stochastic complex systems.
+Brownian motion, originally studied by the botanist Robert Brown in the 19th century and popularized by Einstein in the early 20th century, describes motion that behaves stochastically.
+For instance, if a particle suspended in water were acted upon randomly by other particles, its behavior would be described as Brownian.
+My research with Dr.
+Motsch will seek to demonstrate that Brownian motion will accurately model stochastically complex systems and allow us to invent frameworks that derive meaningful insights into the future.\\
+\indent In order to quantify the accuracy of our approach, we will be using a tool in mathematics called the Wasserstein Distance, which quantifies the distance between functions or data sets, and thereby provides us insight into the improvement that our approach to modeling stochastic systems offers.
+We aim to collaborate with
+Dr.
+Pedro Lowenstein, a professor of Neurosurgery at the University of Michigan to obtain data on cancer growth in a mouse, and with this data, we will use the Wasserstein Distance to compute the distance between the true biological data, the predictions made by the Brownian motion approach, and the predictions made by the deterministic approach.
+We seek to understand how the number of particles used in the Brownian motion approach affects the accuracy of our approximation of the biological data.
+More specifically, we will propose an algorithm that will determine the number of particles necessary to get within a desired error using the Brownian motion approach.
+\\
+\indent During a summer research project, completed under Dr.
+Anne Gelb in 2014, I studied the use of statistical algorithms to improve Fourier edge detection in the presence of corrupted data.
+Specifically, Dr.
+Gelb and I derived an algorithm for reconstructing MRI images that is 166$\%$ more accurate than previous methods available to the medical community.
+By working with Dr.
+Gelb, I developed an interest in using statistical tools to study relevant problems in biology.
+During the summer of 2016 I participated in two summer research programs that provided me with the interest and background necessary to successfully complete my Origins Project research.
+My first project was at the University of California, Santa Barbara where I took a series of courses on differential equations in random media, biological modeling of living systems, and statistical approximations of biological processes.
+Through conversations with professors and graduate students from Princeton, Stanford, UC Berkeley and others, I learned about cutting-edge research at the intersection of biology and statistics and will apply these techniques in my Origins Project research.\\
+\indent In my second project, I was selected as one of three from over 300 undergraduates to participate in a research experience at San Diego State University.
+There I studied the convergence behavior of statistical algorithms used in linear regression models.
+Specifically, my advisor and I proposed the first algorithm for concrete convergence bounds for the error of a Gibbs sampler in a Bayesian linear regression model.
+These results will be submitted for publication in October.
+This research experience taught me the tools to complete statistical convergence analysis, and I will use similar tools to study the convergence behavior of the approximation to complex systems using Brownian motion.
+\\
+\indent To successfully complete this research, I will need an advanced grasp of real analysis, probability theory, and numerical analysis.
+.
+I have taken a course on numerical methods with Dr.
+Motsch, and I have enrolled in graduate courses in real analysis, distribution theory, and stochastic processes.
+In these courses I will obtain the tools necessary to analyze papers relating to my work with Dr Motsch, as well as make meaningful contributions to the field.
+In the Spring I will finish the graduate sequence in real analysis and probability theory, creating a solid theoretical foundation that I will use in my research.\\
+\indent Developing accurate statistical algorithms for understanding stochastic complex systems is one of the largest problems in modern biological sciences, and the opportunity to study this type of problem has been my main focus during my undergraduate career.
+In addition to contributing to the field of computational biology, I hope to motivate theoretical mathematicians and statisticians to solve problems of interest in applied fields, leading to more interactions and discussions between the previously distant groups of researchers.
+Indeed, the Wasserstein distance might also be successfully applied to problems in computer science and chemistry.
+In addition to significantly adding to my knowledge in statistics and computational mathematics, I hope this research project will start a larger discourse on the use of statistical algorithms, specifically Brownian motion, in biology, chemistry, and other applied fields.
+\\
+\indent The Origins Project represents an avenue towards galvanizing an interest among pure mathematicians and statisticians in using theoretical tools to study problems of interest in biology.
+With the Origins Project’s Undergraduate Research Scholarship, I will have the opportunity to demonstrate the utility of mathematics and statistics in computational biology.
+By improving the accuracy of approximations in biology, which studies the development of infectious diseases and our place in the universe, my research will shed light on the origin and future of humanity.
 
- Dr.\ Motsch and I are currently deriving an alternative to the traditional approach to modeling the behavior of biological systems.
-The traditional approach is to use differential equations to model such systems.
-For instance,
-a biologist interested in modeling cancer growth would try to solve:
+\newpage 
+Modeling complex systems, from the growth of cancer cells and bacteria to the movement of stars and planets, is critical to our understanding of the physical world and our place in it.
+Scientists studying the spread of infectious diseases must understand the virus's behavior to effectively treat patients, and by accurately modeling evolutionary dynamics we are able to grasp our origins, as well as possibily our future.
+Modeling these processes becomes difficult when the system is stochastic, and new approaches are needed to analyze such models.
+Brownian motion is an active field of research in statistics, but little research has been done to apply such theory to complex systems.
+My Origins Project research will derive accurate, computationally efficient algorithms for modeling random complex systems using Brownian motion.
 
+\begin{figure}
+\centering
+    \begin{subfigure}[t]{0.48\textwidth}
+      \centering
+\begin{tikzpicture}
+\begin{axis}[legend style={font=\fontsize{8}{4}\selectfont,
+        at={(1.18,.95)},
+        anchor=east},
+    xlabel={$x$},
+    ylabel={$f(x)$},
+    grid = both,
+    legend entries = {Differential Equation, Brownian Motion}
+]
+\addplot table 
+{
+  -3.000000000000000   0.004431848411938
+  -2.900000000000000   0.005952532419776
+  -2.800000000000000   0.007915451582980
+  -2.700000000000000   0.010420934814423
+  -2.600000000000000   0.013582969233686
+  -2.500000000000000   0.017528300493569
+  -2.400000000000000   0.022394530294843
+  -2.300000000000000   0.028327037741601
+  -2.200000000000000   0.035474592846231
+  -2.100000000000000   0.043983595980427
+  -2.000000000000000   0.053990966513188
+  -1.900000000000000   0.065615814774677
+  -1.800000000000000   0.078950158300894
+  -1.700000000000000   0.094049077376887
+  -1.600000000000000   0.110920834679456
+  -1.500000000000000   0.129517595665892
+  -1.400000000000000   0.149727465635745
+  -1.300000000000000   0.171368592047807
+  -1.200000000000000   0.194186054983213
+  -1.100000000000000   0.217852177032551
+  -1.000000000000000   0.241970724519143
+  -0.900000000000000   0.266085249898755
+  -0.800000000000000   0.289691552761483
+  -0.700000000000000   0.312253933366761
+  -0.600000000000000   0.333224602891800
+  -0.500000000000000   0.352065326764300
+  -0.400000000000000   0.368270140303323
+  -0.300000000000000   0.381387815460524
+  -0.200000000000000   0.391042693975456
+  -0.100000000000000   0.396952547477012
+                   0   0.398942280401433
+   0.100000000000000   0.396952547477012
+   0.200000000000000   0.391042693975456
+   0.300000000000000   0.381387815460524
+   0.400000000000000   0.368270140303323
+   0.500000000000000   0.352065326764300
+   0.600000000000000   0.333224602891800
+   0.700000000000000   0.312253933366761
+   0.800000000000000   0.289691552761483
+   0.900000000000000   0.266085249898755
+   1.000000000000000   0.241970724519143
+   1.100000000000000   0.217852177032551
+   1.200000000000000   0.194186054983213
+   1.300000000000000   0.171368592047807
+   1.400000000000000   0.149727465635745
+   1.500000000000000   0.129517595665892
+   1.600000000000000   0.110920834679456
+   1.700000000000000   0.094049077376887
+   1.800000000000000   0.078950158300894
+   1.900000000000000   0.065615814774677
+   2.000000000000000   0.053990966513188
+   2.100000000000000   0.043983595980427
+   2.200000000000000   0.035474592846231
+   2.300000000000000   0.028327037741601
+   2.400000000000000   0.022394530294843
+   2.500000000000000   0.017528300493569
+   2.600000000000000   0.013582969233686
+   2.700000000000000   0.010420934814423
+   2.800000000000000   0.007915451582980
+   2.900000000000000   0.005952532419776
+   3.000000000000000   0.004431848411938
+   };
+
+   \addplot table
+   {
+
+  -3.000000000000000   0.009977212516898
+  -2.900000000000000                   0
+  -2.800000000000000   0.009977212516898
+  -2.700000000000000                   0
+  -2.600000000000000   0.009977212516898
+  -2.500000000000000   0.039908850067592
+  -2.400000000000000   0.049886062584491
+  -2.300000000000000   0.029931637550694
+  -2.200000000000000   0.019954425033796
+  -2.100000000000000   0.069840487618287
+  -2.000000000000000   0.049886062584491
+  -1.900000000000000   0.049886062584491
+  -1.800000000000000   0.149658187753472
+  -1.700000000000000   0.139680975236574
+  -1.600000000000000   0.109749337685879
+  -1.500000000000000   0.139680975236574
+  -1.400000000000000   0.119726550202777
+  -1.300000000000000   0.259407525439351
+  -1.200000000000000   0.259407525439351
+  -1.100000000000000   0.209521462854860
+  -1.000000000000000   0.279361950473147
+  -0.900000000000000   0.269384737956249
+  -0.800000000000000   0.329248013057638
+  -0.700000000000000   0.339225225574536
+  -0.600000000000000   0.309293588023842
+  -0.500000000000000   0.399088500675925
+  -0.400000000000000   0.329248013057638
+  -0.300000000000000   0.409065713192823
+  -0.200000000000000   0.339225225574536
+  -0.100000000000000   0.349202438091434
+                   0   0.379134075642128
+   0.100000000000000   0.478906200811110
+   0.200000000000000   0.438997350743517
+   0.300000000000000   0.369156863125230
+   0.400000000000000   0.289339162990045
+   0.500000000000000   0.349202438091434
+   0.600000000000000   0.249430312922453
+   0.700000000000000   0.289339162990045
+   0.800000000000000   0.239453100405555
+   0.900000000000000   0.359179650608332
+   1.000000000000000   0.199544250337962
+   1.100000000000000   0.209521462854860
+   1.200000000000000   0.169612612787268
+   1.300000000000000   0.139680975236574
+   1.400000000000000   0.099772125168981
+   1.500000000000000   0.149658187753472
+   1.600000000000000   0.049886062584491
+   1.700000000000000   0.109749337685879
+   1.800000000000000   0.049886062584491
+   1.900000000000000   0.109749337685879
+   2.000000000000000   0.049886062584491
+   2.100000000000000   0.009977212516898
+   2.200000000000000   0.059863275101389
+   2.300000000000000                   0
+   2.400000000000000   0.019954425033796
+   2.500000000000000   0.009977212516898
+   2.600000000000000                   0
+   2.700000000000000   0.009977212516898
+   2.800000000000000   0.009977212516898
+   2.900000000000000                   0
+   3.000000000000000                   0
+   };
+\end{axis}
+\end{tikzpicture}
+      \caption{Comparing Approaches}\label{fig:1a}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.48\textwidth}
+      \centering
+\begin{tikzpicture}
+\begin{axis}[legend style={font=\fontsize{15}{5}\selectfont,
+        at={(1.15,0.15)},
+        anchor=east},
+    xlabel={$t$},
+    ylabel={$WD$},
+    grid = both,
+    legend entries = {Wasserstein Distance}
+]
+\addplot table 
+{
+                   0   0.095459745805488 
+   0.500000000000000   0.132619925678194
+   1.000000000000000   0.142863724437454
+   1.500000000000000   0.160699344683304
+   2.000000000000000   0.153470780324526
+   2.500000000000000   0.163047119266944
+   3.000000000000000   0.175466336257786
+   3.500000000000000   0.183119699694153
+   4.000000000000000   0.208128448602138
+   4.500000000000000   0.227184509337919
+   5.000000000000000   0.259384321795910
+   5.500000000000000   0.274999121877102
+   6.000000000000000   0.302427937353585
+   6.500000000000000   0.322783604585681
+   7.000000000000000   0.353867968730460
+   7.500000000000000   0.379632470853626
+   8.000000000000000   0.385074905130309
+   8.500000000000000   0.410417535578750
+   9.000000000000000   0.435227246829623
+   9.500000000000000   0.439715247612350
+  10.000000000000000   0.457522713465409
+};
+
+\end{axis}
+\end{tikzpicture}
+      \caption{Computing the WD}\label{fig:1b}
+    \end{subfigure}
+  \caption{}\label{fig:1}
+\end{figure}
+%
+
+The standard approach to modeling complex systems is to use a differential equation. For instance, a biologist interested in studying cancer growth would try to numerically solve
 \begin{equation}
-\frac{\partial f}{\partial t} = D\Big(\frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}\Big)+ f(1-f).
+\frac{\partial f}{\partial t} = \Big(\frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}\Big)+ f(1-f).
 \label{eq: Macro}
@@
 
-To solve such an equation, one discretizes the physical space (corresponding to the $x$ and $y$ variables) and the temporal space (corresponding to the $t$ variable).
-This approach requires the ratio $\frac{\Delta t}{\Delta x^2}$ to be very small in order for the solution to be accurate.
-This results in a computationally expensive problem.
-Furthermore, as the physical space gets larger, the computational time required increases exponentially.
-For problems of even moderate size, this approach is therefore impracticle.\\
-
-\begin{wrapfigure}{L}{0.4\textwidth}
-  \begin{center}
-   \includegraphics[width=0.4\textwidth]{OriginsPic.jpg}
-  \end{center}
-  \caption{Comparison of Approaches}
-\end{wrapfigure}
-
-\noindent The second approach is to model the behavior of the physical system using particles.
-Instead of solving a complicated differential equation, one begins with a collection of particles, with $X_i^{m}$ denoting the $i$-th particle at time $m.$ One then computes the position of these particles at a future time with
+\noindent While this approach is common, it fails to accurately model the stochastic components of the systems.
+Dr.\ Motsch and I will examine this same system through the use of Brownian motion.
+Brownian motion, originally studied by the botanist Robert Brown in the 19th century and popularized by Einstein in the early 20th century, describes motion that behaves in a random way.
+If an variable's motion is Brownian, its derivative is a normally distributed random variable.
+By letting $X^m$ denote the particle's location at time $m$, one can use the following equation to compute its future location
 \begin{equation}
-x_i^{m+1} = x_i^{m} + C\textbf{W},
+X^{m+1} = X^{m} + C\textbf{W},
 \label{eq: Micro}
 \end{equation}
-where $C$ is a constant and $\textbf{W} \sim \mathcal{N}(0, 1).$ One simply needs to draw normally distributed numbers in order to approximate the behavior of the system.
-One can prove mathematically that these particles converge in distribution to the equation of the differential equation.
+
+\noindent where $C$ is a constant and $\textbf{W} \sim \mathcal{N}(0, 1).$ Substantial research has been done on theoretical Brownian motion, with little progress on applying these results to complex systems.
+
+The goal of my research under Dr.
+Motsch will be to demonstrate that modeling complex systems with Brownian motion, rather than with a differential equation, leads to a more accurate approximation of the underlying dynamics.
+We will study a reaction-diffusion model in one and two dimensions that can be used to study the growth and death of cancer cells in an organism.
+In this model, cells are created and destroyed randomly, and so we believe modeling this system using a deterministic differential equation leads to incaccurate solutions.
+This model on its own is used frequently in academia and industry to understand the behavior of cancer and infectious diseases, and thus an improved solution to this important model will have substantial impact on theoretical and computational biology.
+To show that the Brownian motion approach is more accutarate, we will compare our results to real-world data obtained from an ASU biology lab.
+In order to make these comparisons, we will be using a tool in pure mathematics called the Wasserstein Distance (WD), which measures the difference between two data sets.
+The WD will enable us to quantify the improvement of the Brownian motion approach.
+
+Dr.\ Motsch and I are well positioned to complete this project.
+I am an undergraduate student working in applied mathematics who intends to pursue a PhD in statistics.
+In particular I am interested in applying statistical techniques to solving real-world problems.
+This past summer I was selected along with three other undergraduates out of over 300 to participate in a research program at San Diego State University.
+There my advisor and I derived the first method of obtaining convergence behavior for the error in algorithms that one uses when estimating coefficients in the linear regression equation.
+This project resulted in a publication, and I developed an interest in applying statistical tools to study real-world problems.
+My role in the project will be to appply statistical techniques to modeling Brownian motion in one and two dimensions and applying numerical techniques to solving the differential equations we'll be using.
+Dr.\ Motsch received his PhD in applied mathematics with an emphasis in animal cognition and mathematical biology, and works at the intersection of computational math, biology, and statistics.
+He has overseen undergraduate and graduate theses in flocking behavior, applied statistics, and offers the biological perspective necessary for this project.
+He has worked previously in the Wasserstein Distance and will provide the background in the WD to apply it to our research.
+
+This project requires an advanced grasp on real analysis, probability theory, and numerical analysis.
+In order to complete this project successfully, I have enrolled in graduate courses in real analysis,  distribution theory, and stochastic processes.
+In these courses I will obtain the tools necessary to read papers relating to my project as well as hold informed discussions with Dr.
+Motsch on our research.
+I have taken a course on numerical methods with Dr.\ Motsch, and I will use my textbook as a reference during my research.
+In the Spring I will finish the graduate sequence in real analysis and probability theory.
+
+The Origins Project Scholarship will allow me to focus solely on my research with Dr.\ Motsch, instead of maintaining an on-campus job to pay for college.
+This will allow me to make faster progress on my research.
+Having this research experience will also give me the experience necessary to perform well as a graduate student.
+We believe that the underlying dynamics of stochastic complex systems are different in nature from deterministic models and are therefore not accurately modeled by differential equations.
+Through this project I seek to better understand these dynamics through statistical modeling.
+My research with Dr.
+\ Motsch will demonstrate the possibility of using Brownian motion to model stochastic complex systems, shedding light on the random forces that determine the physical world and on our role as a species in it.
+
+ \newpage
+
+Through this research I will gain the necessary research experience and technical background to succesfully enter into a PhD program in statistics.
+By publishing our results, I will have an advantage in the application process for graduate school.
+By conducting this research under Dr.\ Motsch, I'll gain
+
+This project allows me to apply statistical and numerical techniques to solve tangible problems and provides me with an ideal transition to graduate school.
+We will be using techniques such as Brownian motion,  approximations of intractable distributions, finite-difference techniques to study differential equations.
+lysis and graduate distribution theory in the fall to build upon my theoretical foundation.
+
+ Dr.\ Motsch received his PhD in applied mathematics with an emphasis in animal cognition and mathematical biology, and works at the intersection of computational math and biology.
+He has mentored several undergraduates working in computational biology, flocking patterns, and cell behavior.
+
+The outline of the project is as follows: In September and October, Dr.\ Motsch and I will be submitting a paper to journal that demonstrates that agent-based modeling of complex systems accurately represents the behavior of the complex system using data from \textbf{Motsch, who is this data from?}.
+In November and December, we will study  and finally in the Spring we will examine.
+
+As a recipient of the Origins Project Scholarship, I will have the chance to devote myself to my research project, rather than working to pay for college, allowing me and Dr.
+\ Motsch to make quick progress in our research.
+
+The Origins Project Scholarship will allow me to focus solely on my research with Dr.\ Motsch, instead of working on campus, allowing for more substantial research progress during the academic year.
+This will in turn allow me to prepare a stronger graduate school application,
+ The Origins network will provide me and Dr.\ Motsch with an invaluable network of researchers in biology, chemistry and other fields we can speak with, collaborate with, and learn from.
+This will in turn lead to research that understands the needs of these fields, leading to more helpful and effective results.
+
+\newpage 
+\begin{center}
+\textbf{Notes}
+\end{center}
+
  In Figure 1,
 the solutions from (\ref{eq: Macro}) and (\ref{eq: Micro}) are plotted.
+
+\textbf{Agent is small, the macro might not be realistic, might not describe. send code and kernel, image I have, and send draft.}
+
 This approach is computationally faster and places no restriction on the physical and temportal discretization, yet there is a problem.
@@
 
+there is a miscmatch, we sometimess need to use micro because it more accurately reflects the underlying truth. agent-based models, complex systems.
+
+Novel Techniques to Modeling Complex Systems
+
+Comparison of Agent-based and   
+agent-based models for complex systems:
+
+Modeling Complex Systems: Agent-based Models vs. Macro Models.
+Contrasting
+
+you need large particle for consistency, we need to shift that the 
+macro works if number is large enough, but sometimes the system is too small  so that the validity of the macro is questionable.
+
+it simplifies the computation and allows for more diverse models to be considered.
+
+two figures: one where it works, one where it doesnt work., the one where that it doesnt work, in some sort of experimental data, we'll hold off for now.
+we want to use opacity, level curves, difference both macro and micro.
+
+open question: compare micro vs. macro, sometimes one doesnt work when n isnt large enough. We need statistics to study the behavior of agent-based when the macro is not enough.
+
+To solve such an equation, one discretizes the physical space (corresponding to the $x$ and $y$ variables) and the temporal space (corresponding to the $t$ variable).
+This approach requires the ratio $\frac{\Delta t}{\Delta x^2}$ to be very small in order for the solution to be accurate.
+This results in a computationally expensive problem.
+Furthermore, as the physical space gets larger, the computational time required increases exponentially.
+For problems of even moderate size, this approach is therefore impracticle.\\
+
+Many biological processes can be
+
+As our computing power inceases, questions that once seemed impossible to analyze become accessible.
+Faster computation has faciliatated new discoveries in computer science, physics, and chemistry, but nowhere are these discoveries more important to our understanding of the physical world than in biology.
+Biological processes, such as the growth of cancer, the spread of disease, or the evolution of a species, are at the forefront of modern science, and understanding these processes is critical to understanding our place in the universe.
+Working under Dr.
+S\'{e}bastien Motsch, my Origins Project work will focus on deriving faster, more accurate approaches to analyzing complex dynamical systems.
+
 \end{document}
+
+                  0   0.095459745805488 
+   0.100000000000000   0.103038153835043
+   0.200000000000000   0.122390116625128
+   0.300000000000000   0.132687398532943
+   0.400000000000000   0.141253013593047
+   0.500000000000000   0.132619925678194
+   0.600000000000000   0.132080641788240
+   0.700000000000000   0.126684962315643
+   0.800000000000000   0.126916579284260
+   0.900000000000000   0.130572856315571
+   1.000000000000000   0.142863724437454
+   1.100000000000000   0.151314041878340
+   1.200000000000000   0.157724124185497
+   1.300000000000000   0.157487782560854
+   1.400000000000000   0.150699563747470
+   1.500000000000000   0.160699344683304
+   1.600000000000000   0.138550211752228
+   1.700000000000000   0.141804472307241
+   1.800000000000000   0.139818367972234
+   1.900000000000000   0.146181574882163
+   2.000000000000000   0.153470780324526
+   2.100000000000000   0.157789500372874
+   2.200000000000000   0.164623391004034
+   2.300000000000000   0.174651856775786
+   2.400000000000000   0.165106253910355
+   2.500000000000000   0.163047119266944
+   2.600000000000000   0.165371933076515
+   2.700000000000000   0.168312824055832
+   2.800000000000000   0.163027901066160
+   2.900000000000000   0.165021445928958
+   3.000000000000000   0.175466336257786
+   3.100000000000000   0.169441333113976
+   3.200000000000000   0.167736362245388
+   3.300000000000000   0.174500263853999
+   3.400000000000000   0.176807926512032
+   3.500000000000000   0.183119699694153
+   3.600000000000000   0.186685386248164
+   3.700000000000000   0.190381205104428
+   3.800000000000000   0.196700741959594
+   3.900000000000000   0.199003478468923
+   4.000000000000000   0.208128448602138
+   4.100000000000001   0.212927381498538
+   4.200000000000000   0.216035775125143
+   4.300000000000000   0.216309196519999
+   4.400000000000000   0.226192482106804
+   4.500000000000000   0.227184509337919
+   4.600000000000001   0.232672773797666
+   4.700000000000000   0.244151932180668
+   4.800000000000001   0.245019748714593
+   4.900000000000000   0.247101950208486
+   5.000000000000000   0.259384321795910
+   5.100000000000000   0.262244593436841
+   5.199999999999999   0.263736387029788
+   5.300000000000000   0.271223425250633
+   5.399999999999999   0.269171050595840
+   5.500000000000000   0.274999121877102
+   5.600000000000000   0.280189726935880
+   5.700000000000000   0.278080380840482
+   5.800000000000000   0.284123569488352
+   5.899999999999999   0.293092319792321
+   6.000000000000000   0.302427937353585
+   6.100000000000000   0.304988685584145
+   6.199999999999999   0.310875304199478
+   6.300000000000000   0.309492394889520
+   6.400000000000000   0.316683383435123
+   6.500000000000000   0.322783604585681
+   6.600000000000000   0.327349943001851
+   6.699999999999999   0.337347746399912
+   6.800000000000000   0.344809074579488
+   6.900000000000000   0.350911235743597
+   7.000000000000000   0.353867968730460
+   7.100000000000000   0.359376409561387
+   7.199999999999999   0.367885437481766
+   7.300000000000000   0.372205234477801
+   7.400000000000000   0.376966998227070
+   7.500000000000000   0.379632470853626
+   7.600000000000000   0.382266518732972
+   7.699999999999999   0.379917749569585
+   7.800000000000000   0.378863138190330
+   7.900000000000000   0.386603623292818
+   8.000000000000000   0.385074905130309
+   8.100000000000000   0.392721329717624
+   8.199999999999999   0.395608310610736
+   8.300000000000001   0.399023272131075
+   8.400000000000000   0.406569688035700
+   8.500000000000000   0.410417535578750
+   8.600000000000000   0.419335508882418
+   8.699999999999999   0.423990982698414
+   8.800000000000001   0.424340729563118
+   8.900000000000000   0.423686816264073
+   9.000000000000000   0.435227246829623
+   9.100000000000000   0.435876153253802
+   9.199999999999999   0.428210125789563
+   9.300000000000001   0.432099295650654
+   9.400000000000000   0.438695163168372
+   9.500000000000000   0.439715247612350
+   9.600000000000000   0.442966420891081
+   9.699999999999999   0.444992993534448
+   9.800000000000001   0.453284818856704
+   9.900000000000000   0.459222943838312
+  10.000000000000000   0.457522713465409
+
+  Cancer researchers rely on accurate approximations of the cell's behavior to derive more effective treatments, and
+
+What we dont understand in biology can be modeled by stochastic components, and we believe that this research will lead to breakthroughs in fields of biology
+
+-there is more going on than what is reflected in the deterministic. 
+- pde occurs becaose of LLN, but here we are in the case of when we are not in this case.
+
+ has two main goals.
+The first goal is to more accurately model random complex systems using Brownian motion.
+By using tools in real analysis, probability theory, and numerical methods, our work will seek to show that Brownian motion can better model complex systems than the current approach.
+The second goal of our project is to quantify the improvement our approach has over the current approach.
+
+In order to compare the accuracy of our approach to the accuracy of the traditional approach to modeling complex system, we need a way to quantify the difference between our solutions.
+In the past few years a new tools was proposed.
+much research has been done on the Wasserstein Distance, a tool in pure mathematics that quantifies the difference between functions.
+Entirely theoretical, we want to show that this distance can be used quite effectively to quantify the difference between the Brownian motion approach and the differential equation approach.
+The second goal of this project is to take a tool that has been up until now a purely theoretical tool and show that it can be used to effectively quantify the difference in approaches to modeling stochastic complex systems.

### 2016-10-18 TestLocDIff1D.m

@@ -1,0 +1,17 @@
+% This code solves the diffusion equation in one dimension using a
+% diffusion coefficient that depends on the the physical location. 
+
+x = linspace(-3,3,10);
+t = linspace(0,1,.1);
+D = linspace(0,5,numel(x));
+rho = zeros(numel(x), numel(t));
+deltaX = x(2)-x(1);
+
+rho(:,1) = normpdf(x,0,1);
+
+
+for i = 2:numel(t)
+    for j = 2:numel(x)
+       rho(
+    end
+end

### 2016-10-18 model_1D/Micro_Macro.m

@@ -9,13 +9,14 @@
 % Parameters model
 D = 2/3;
 
+
 % Numerical parameters
 dt = .1;
-T  = 2;
+T  = 10;
 dx = .1;
 L  = 3;
 method_Kpp = 2;                           % 1: explicit, 2: Cranck-Nicholson
-Ninit  = 1e2;                           % number of particles initially
+Ninit  = 10^3                           % number of particles initially
 seedNum = 3;
 % Initial condition
 x = -L:dx:L;
@@ -37,10 +38,12 @@ rhoMicro = KPP_Micro(rhoIC,x,D,dt,T,Ninit,seedNum);
 %% B.1) Wasserstein distance
 addpath('lib')
 nT = length(rhoMacro(1,:));
-stock_WD = zeros(1,nT);
+%stock_WD = zeros(1,nT);
+
 for k = 1:nT
-    stock_WD(k) = WD_Cont(rhoMicro(:,k),rhoMacro(:,k),x);
+    stock_WD1(k) = WD_Cont(rhoMicro(:,k),rhoMacro(:,k),x);
 end
+
 %% B.2) Velocity traveling wave
 xStar_micro = zeros(1,nT);
 xStar_macro = zeros(1,nT);
@@ -92,3 +95,6 @@ coeff = polyfit(intT,xStar_micro,1);
 speedMicro = coeff(1);
 coeff = polyfit(intT,xStar_macro,1);
 speedMacro = coeff(1);
+
+
+end

### 2016-10-18 model_2D/lib/ComputeVelocity.m

@@ -1,7 +1,20 @@
-function distance = ComputeVelocity(x,y,rhoMacro,c)
+function f_int = ComputeVelocity(x,y,rho,r)
 
-[Y,I] = min(abs(rhoMacro(:)-c));
-[row, col] = ind2sub(size(rhoMacro),I);
-distance = sqrt(x(row)^2 + y(col)^2);
+xRestrict = zeros(length(x),1); yRestrict = zeros(length(y),1);
+
+for i = 1:length(r)
+for k = 1:length(y)
+    for j = 1:length(x)
+        if (x(j).^2+y(k).^2 <= r(i).^2)
+            xRestrict(j) = 1; yRestrict(k) = 1;
+        end
+    end    
+end
+
+f_res = rho(logical(xRestrict),logical(yRestrict),1);
+f_int(i) = 1/(2*pi*r(i)) * trapz(x(find(xRestrict)),trapz(y(find(yRestrict)),f_res));
+
+end
+
 
 end

### 2016-10-18 model_2D/lib/Macro.m

@@ -26,9 +26,6 @@ function [rhoDir,rho] = Macro(x,y,D,dt,T,mu1,mu2,var1,var2,shouldPlot)
     dy = y(2)-y(1);
     nX = length(x);
     nY = length(y);
-    T = 20;
-    dt = 10^-2;
-    D = 1;
     nT = floor(T/dt + .5);
    
     [X1,X2] = meshgrid(x,y);

### 2016-10-18 model_2D/lib/Micro.m

@@ -1,14 +1,20 @@
-     function rho = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2)
+function [rho,X] = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2)
+
 % 
 % Solve the KPP equation as a particle system:
 %    dX_t = ?dB_t
 %  + 'birth process' with intensity 1-?.
 % 
-
+    [X1,X2] = meshgrid(x,y);
+    X1_tp = X1';
+    X2_tp = X2';
+        nX = length(x);
+    nY = length(y);
     [xmesh,ymesh] = meshgrid(x,y);
     rho = zeros(numel(x),numel(y),floor(T/dt + .5));
     
     X = [mu1 + sqrt(var1)*randn(N,1), mu2 + sqrt(var2)*randn(N,1)];
+    %rho(:,:,1) = reshape(mvnpdf([X1_tp(:) X2_tp(:)],[mu1, mu2],[var1, var2]),nX,nY); 
     rho(:,:,1) = hist3(X,[numel(x),numel(y)]);
    
     for j = 2:floor(T/dt + .5)
@@ -18,7 +24,9 @@
         X = X + sqrt(2*D)*sqrt(dt)*randn(M,2); 
             
         % Histogram
-        rho(:,:,j) = hist3(X,[numel(x),numel(y)]); 
+        %rho(:,:,j) = hist3(X,[numel(x),numel(y)]); 
+        [bandwidth, density, x_vec, y_vec] = kde2d(X,numel(x));
+        rho(:,:,j) = density;
         [Xtemp1, Xtemp2] = meshgrid(X(:,1), X(:,2));
         rho_at_X = interp2(xmesh,ymesh,rho(:,:,j),Xtemp1(:,1),Xtemp2(:,1)); % interpolate at x_i, coming from Brownian paths.
         

### 2016-10-18 model_2D/lib/OptimMatrices.m

@@ -1,0 +1,38 @@
+function [A, b, C ] = OptimMatrices(f,g,x,y)
+
+    [X1, X2] = meshgrid(x,y);
+    XMESH = [X1(:) X2(:)];
+    Xpos = XMESH;
+    Ypos = Xpos;
+    nX = length(Xpos);
+    nY = length(Ypos);
+
+    b = [f(:); g(:)];
+    
+    A = zeros(nX+nY,nX*nY);
+    for k=1:nX
+        A(k, (k-1)*nY+[1:nY]) = 1;
+    end
+    for k=1:nY
+        for k2=1:nX
+            A(k+nX, k + (k2-1)*nY) = 1;
+        end
+    end
+    
+%      % We have an extra constraint (last equality is automatically satisfied)
+      A(end,:) = [];
+      b(end)   = [];
+      
+          nX = length(Xpos(:,1));
+    nY = length(Ypos(:,1));
+    matrix_C = zeros(nX,nY);
+    for i=1:nX
+        for j=1:nY
+            matrix_C(i,j) = sqrt( (Ypos(j,1)-Xpos(i,1))^2 + (Ypos(j,2)-Xpos(i,2))^2 );
+        end
+    end
+    C = matrix_C(:);
+
+
+end
+

### 2016-10-18 model_2D/lib/TwoDStochastic.m

@@ -6,19 +6,21 @@
 clc; clear all; close all;
 
 %% Parameters
-D = 5;                                % diffusion coefficients
+D = 1;                                % diffusion coefficients
 %% Parameters micro
-N = 10^4;                               % number of cells (initially)
+N = 10^2;                               % number of cells (initially)
 Mass_rhoIC = 2;
 seedNum = 1;
 % domain
-dx = .5;
-dy = .5;
-x = -5:dx:5;
-y = -5:dy:5;
+dx = .1;
+dy = .1;
+%x = -5:dx:5;
+x = linspace(-5,5,2^3);
+y = x;
+%y = -5:dy:5;
 % time
-dt = .1;
-T =   5;
+dt = .01;
+T =   1;
 t = 0:dt:T;
 
 % Initial Conditiion
@@ -27,10 +29,18 @@ mu2 = 0;
 var1 = 1;
 var2 = 1;
 shouldPlot = 0;
+r = linspace(1.5,40,100);
+VelMac = zeros(length(t),length(r));
+VelMic = zeros(length(t),length(r));
 
 
-rhoMicro = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2);
+
+
+[rhoMicro, Part] = Micro(x,y,D,dt,T,N,mu1,mu2,var1,var2);
 [rhoDir,rho] = Macro(x,y,D,dt,T,mu1,mu2,var1,var2,true);
+for l = 1:length(t)
+   VelMac(l,:) = ComputeVelocity(x,y,rho(:,:,l),r); 
+end
  
 
 for q = 1:length(t)
@@ -59,45 +69,8 @@ rho(:,:,q) = rho(:,:,q)/sum(sum(rho(:,:,q)));
 %[WD(q),Diff1,Diff2,C,A,b] = TwoDOptimizationCode(x,y,rhoMicro(:,:,q),...
  %                                     rho(:,:,q));
 %SimplexCode = toc;
-
-f = rho(:,:,q);
-g = rhoMicro(:,:,q);
-
-    [X1, X2] = meshgrid(x,y);
-    XMESH = [X1(:) X2(:)];
-    Xpos = XMESH;
-    Ypos = Xpos;
-    nX = length(Xpos);
-    nY = length(Ypos);
-
-    b = [f(:); g(:)];
-    
-    A = zeros(nX+nY,nX*nY);
-    for k=1:nX
-        A(k, (k-1)*nY+[1:nY]) = 1;
-    end
-    for k=1:nY
-        for k2=1:nX
[…]

### 2016-10-18 model_2D/lib/radial_density.m

@@ -1,0 +1,68 @@
+function [densityMic,intR,int_rhoR] = radial_density(x,y,rhoMicro,rhoMacro,Part)
+
+% Parameters
+dr = .05;
+intR = 0:dr:5;
+
+% Initialize mean vectors
+x_barMac = zeros(1,2); 
+x_barMic = zeros(1,2);
+rhoTempX = zeros(length(x),length(y));
+rhoTempY = zeros(length(x),length(y));
+
+for k = 1:length(rhoMicro(1,:,1))
+rhoTempX(:,k) = x.*rhoMacro(k,:);
+end
+
+for k = 1:length(rhoMacro(:,1,1))    
+rhoTempY(k,:) = y.*rhoMacro(:,k)';
+end
+
+% Compute x_bar
+x_barMac(1) = 1/(trapz(x,trapz(y,rhoMacro))) * trapz(x,trapz(y,rhoTempX));
+x_barMac(2) = 1/(trapz(x,trapz(y,rhoMacro))) * trapz(x,trapz(y,rhoTempY));
+
+f = @(x,y) (x.^2+y.^2<1);
+Mesh_x_barMac1 = x_barMac(1)*ones(length(x),length(x));
+Mesh_x_barMac2 = x_barMac(2) *ones(length(y),length(y));
+indicator_DiscR = @(x,y,R) ((x - Mesh_x_barMac1).^2 + (y - Mesh_x_barMac2).^2 < R^2);
+
+% init
+[X,Y] = meshgrid(x,y);
+%A = f(X,Y);
+A = exp(-X.^2-Y.^2);
+nR = length(intR);
+int_rhoR = zeros(1,nR);
+
+
+% loop
+for k=1:nR
+    % init
+    r = intR(k);
+    B = indicator_DiscR(X,Y,r);
+    % the value
+    int_rhoR(k) = trapz(x,trapz(y,A.*B));
+end
+
+% plot
+% intR_mid = (dr/2):dr:(2-dr/2);
+% rhoR = diff(int_rhoR)./(2*pi*intR_mid*dr);
+% grid on;
+
+% plot(intR_mid,rhoR, intR_mid,exp(-intR_mid.^2))
+% xlabel('r')
+
+% Micro
+x_barMic(1) = mean(Part(:,1));
+x_barMic(2) = mean(Part(:,2));
+
+for k=1:nR
+    % init
+    r = intR(k);
[…]
